---
title: "Comparing CellChat and CellPhoneDB results"
output: github_document
date: "2025-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Crosscheck function

To proceed with the crosscheck function, users should have ran both `run_cellchat()` and `run_cellphonedb()` on their Seurat object. If these steps have not been performed, refer to tutorials to [run CellChat](https://github.com/nigelhojinker/CCCtools/blob/main/data-raw/Run_CellChat.md) and [run CellPhoneDB](https://github.com/nigelhojinker/CCCtools/blob/main/data-raw/Run_CellPhoneDB.md) at the respective links. 

We provide a function `crosscheck()` that takes as input both the CellChat object from `run_cellchat()` and the list from `run_cellphonedb()`, as well as an optional `threshold` argument should users wish to adjust the p-value (default: 0.05). `crosscheck()` returns a list of two elements: (1) Comparative analysis result (2) 3x3 contingency table of interaction types in both CellChat and CellPhoneDB. We categorise interactions in the filtered result in terms of the agreement (and lack thereof) between CellChat and CellPhoneDB as such:

- Sig_Both: Interaction is significant in both CellChat and CellPhoneDB
- Sig_CellChat_Not_Found_CellPhoneDB: Interaction is found only in CellChat, and p-value < threshold
- Sig_CellPhoneDB_Not_Found_CellChat: Interaction is found only in CellPhoneDB, and p-value < threshold
- Sig_CellChat_Not_Sig_CellPhoneDB: Interaction is significant in CellChat but not significant in CellPhoneDB
- Sig_CellPhoneDB_Not_Sig_CellChat: Interaction is significant in CellPhoneDB but not significant in CellChat

```{r, eval = FALSE}
data(seu.NL)

cellchat <- run_cellchat(seu.NL, group.by = "labels")

## After setting up python env
cpdb <- run_cellphonedb(seu.NL, labels = "labels")

combine <- crosscheck(cellchat = cellchat, cellphonedb = cpdb)
# Joining with `by = join_by(interaction_name)`
# Joining with `by = join_by(id_cp_interaction, classification)`
# Comparative analysis done successfully for p-value 0.05 
#  CCCtools summary: 
#            CellPhoneDB
# CellChat      Sig Not_Sig Not_Found   Sum
#   Sig         220      23       470   713
#   Not_Sig     705   31748     29763 62216
#   Not_Found   585   21295         0 21880
#   Sum        1510   53066     30233 84809

combine %>% names()
# [1] "result"       "summary"
```
