---
title: "Database customisation by scpeakeR - Part 2"
author: "Ho Jin Ker Nigel, Adaikalavan Ramasamy, Gokce Oguz"
description: >
  Database customisation of CellChatDB.human(v2.2.0) and CellPhoneDB (V5.0) with the proposed changes in Part 1
date: 2025/01/05
format: 
  html:
    toc: true
    toc-location: left
    self-contained: true
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

# Customising CCDB and CPDB

In this document, we implement the corrections identified in "Database harmonisation by scpeakeR - Part 1". If you have not read Part 1, you may proceed to that document [here](https://nigelhojinker.github.io/scpeakeR/data-raw/Database_harmonisation_part1.html).

## Load required packages

```{r}
pacman::p_load(tidyverse, janitor, CellChat, reticulate)
setwd(this.path::here())
rm(list = ls())
```

## Load saved R objects from Part 1 Database harmonisation

```{r}
load("../data/db_utils/CC_int.rda")
load("../data/db_utils/CPDB_int.rda")
load("../data/db_utils/flipped.rda")
```

## CellChatDB -- new

To customise CCDB for cell-cell communication analysis with CellChat, we refer to the document attached [here](https://rdrr.io/github/sqjin/CellChat/f/tutorial/Update-CellChatDB.Rmd).

CellChatDB.human is a list of four elements encompassing the human interactions curated by CellChat package:

```{r}
CellChatDB.human %>% names()
```

We manually added our changes to the individual elements of this database:

### CellChatDB.human\$interaction

```{r}
interaction_input <- CellChatDB.human$interaction
interaction_input$interaction_name <- CC.int$interaction_name
rownames(interaction_input) <- interaction_input$interaction_name
```

### CellChatDB.human\$complex

```{r}
complex_input <- CellChatDB.human$complex %>% 
  rownames_to_column("CellChat_complex") %>% 
  mutate(subunit_1 = ifelse(grepl("^CALCRL", CellChat_complex), "CALCRL", subunit_1),
         subunit_2 = case_when(CellChat_complex == "RA-ALDH1A1" ~ "",
                               CellChat_complex == "Ach-CHAT_SLC10A4" ~ "CHAT",
                               .default = subunit_2),
         subunit_3 = case_when(CellChat_complex == "Ach-CHAT_SLC10A4" ~ "",
                               CellChat_complex == "CALCRL_RAMP3" ~ "GPER1",
                               .default = subunit_3)) %>% 
  column_to_rownames("CellChat_complex")
```

### CellChatDB.human\$cofactor

```{r}
# cofactor slot is unchanged
cofactor_input <- CellChatDB.human$cofactor
```

### CellChatDB.human\$geneInfo

We extract interaction partners of simple genes and add the new changes to geneInfo slot

```{r}
CC.ligand.simple   <- CC.int %>% filter(! complex_ligand)   %>% select(Symbol = ligand_name,   uniprot = ligand)
CC.receptor.simple <- CC.int %>% filter(! complex_receptor) %>% select(Symbol = receptor_name, uniprot = receptor)

CC.gene.map <- bind_rows(CC.ligand.simple, CC.receptor.simple) %>% distinct()
```

```{r}
CC.gene.map %>% dim()
CC.gene.map %>% head()
```

```{r}
geneInfo <- CellChatDB.human$geneInfo %>% 
  left_join(CC.gene.map) %>% 
  mutate(EntryID.uniprot = case_when(is.na(uniprot) ~ EntryID.uniprot,
                                     is.na(EntryID.uniprot) ~ uniprot,
                                     EntryID.uniprot != uniprot ~ uniprot,
                                     .default = EntryID.uniprot)) %>% 
  select(-uniprot)

rm(CC.ligand.simple, CC.receptor.simple, CC.gene.map)
```

## Re-create CellChatDB.human with added changes

```{r}
CellChatDB.human.new <- list()
CellChatDB.human.new$interaction <- interaction_input
CellChatDB.human.new$complex     <- complex_input
CellChatDB.human.new$cofactor    <- cofactor_input
CellChatDB.human.new$geneInfo    <- geneInfo
```

```{r, eval = FALSE}
save(CellChatDB.human.new, file = "../data/CellChatDB.human.new")
```

To summarise, we reconstructed `CellChatDB.human` with our changes from Part 1's database harmonisation and provide `CellChatDB.human.new` for running CellChat analysis. If users only want to perform CellChat analysis on their Seurat object(s), `CellChatDB.human.new` would be the selected database for the run. This allows users to compare CellChat's results against CellPhoneDB's inferred results post-harmonisation.

```{r}
rm(interaction_input, complex_input, geneInfo, cofactor_input, CellChatDB.human.new)
```

## CellPhoneDB -- new

In this section, we implement the same changes after performing database harmonisation to CellPhoneDB. We recreated the database in a format (.zip file) that is readable to CellPhoneDB's internal code. To customise CPDB with our revised changes, we refer to the document [here](https://github.com/ventolab/CellphoneDB/blob/master/notebooks/T0_BuildDBfromFiles.ipynb).

### Load CellPhoneDBv5 input tables for database customisation

```{r, eval = FALSE}
sapply(c("../data", "../data/sources"), dir.create, recursive = TRUE)
```

```{r}
basepath <- "https://raw.githubusercontent.com/ventolab/cellphonedb-data/71ffa8a163389913907d40993bec2821a111f64d/data"

interaction_input <- read_csv(url(file.path(basepath, "interaction_input.csv")))
gene_input        <- read_csv(url(file.path(basepath, "gene_input.csv")))
complex_input     <- read_csv(url(file.path(basepath, "complex_input.csv")))
protein_input     <- read_csv(url(file.path(basepath, "protein_input.csv")))
transcription_factor_input <- read_csv(url(file.path(basepath, "transcription_factor_input.csv")))
uniprot_syn       <- read_tsv(url(file.path(basepath, "sources/uniprot_synonyms.tsv")))
```

### customise interaction_input

```{r}
interaction_input <- interaction_input %>% 
  mutate(partner_a = case_when(partner_a == "HLAA"  ~ "P04439",
                               partner_a == "HLAB"  ~ "P01889",
                               partner_a == "HLAC"  ~ "P10321",
                               partner_a == "IFNA1" ~ "P01562",
                               .default = partner_a),
         partner_b = ifelse(partner_b == "Glycoprotein_Ib_complex", "GP_complex", partner_b))
```

We implement the swapping of interaction partners from Part 1 here

```{r}
to_flip <- which(interaction_input$interactors %in% flipped)

# Make a copy to perform swap
tmp <- interaction_input

interaction_input$partner_a[to_flip] <- tmp$partner_b[to_flip]
interaction_input$partner_b[to_flip] <- tmp$partner_a[to_flip]

interaction_input$protein_name_a[to_flip] <- tmp$protein_name_b[to_flip]
interaction_input$protein_name_b[to_flip] <- tmp$protein_name_a[to_flip]

rm(tmp, flipped, to_flip)
```

### customise gene_input

```{r}
new_genes <- data.frame(gene_name = c("GP9", "GP5"),
                        uniprot = c("P14770", "P40197"),
                        hgnc_symbol = c("GP9", "GP5"),
                        ensembl = c("ENSG00000169704", "ENSG00000178732"))

gene_input <- gene_input %>% 
  mutate(uniprot = case_when(uniprot == "HLAA"  ~ "P04439",
                             uniprot == "HLAB"  ~ "P01889",
                             uniprot == "HLAC"  ~ "P10321",
                             uniprot == "IFNA1" ~ "P01562",
                             .default = uniprot)) %>% 
  bind_rows(new_genes)
```

### customise complex_input

```{r}
complex_input <- complex_input %>% 
  mutate(complex_name = ifelse(complex_name == "Glycoprotein_Ib_complex", "GP_complex", complex_name),
         uniprot_3 = case_when(complex_name == "IL15_receptor" ~ "P31785",
                               complex_name == "TGFbeta_receptor2" ~ "P36897",
                               complex_name == "GP_complex" ~ "P14770",
                               .default = uniprot_3),
         uniprot_4 = ifelse(complex_name == "GP_complex", "P40197", uniprot_4))
```

### customise protein_input

```{r}
new_proteins <- data.frame(uniprot = c("P14770", "P40197"),
                           protein_name = c("GPIX_HUMAN", "GPV_HUMAN"))

protein_input <- protein_input %>% 
  mutate(uniprot = case_when(uniprot == "HLAA"  ~ "P04439",
                             uniprot == "HLAB"  ~ "P01889",
                             uniprot == "HLAC"  ~ "P10321",
                             uniprot == "IFNA1" ~ "P01562",
                             .default = uniprot)) %>% 
  bind_rows(new_proteins)
```

### Save input files to data dir

```{r, eval = FALSE}
write_csv(interaction_input, file = "../data/interaction_input.csv")
write_csv(gene_input,        file = "../data/gene_input.csv")
write_csv(complex_input,     file = "../data/complex_input.csv")
write_csv(protein_input,     file = "../data/protein_input.csv")
write_csv(transcription_factor_input, file = "../data/transcription_factor_input.csv")
write_tsv(uniprot_syn, file = "../data/sources/uniprot_synonyms.tsv")
```

### Creating customised CPDB

```{r, eval = FALSE}
use_condaenv("cpdb")

db_create <- import("cellphonedb.utils.db_utils")
db_create$create_db("../data")
# Created data\cellphonedb_11_07_2025_010457.zip successfully
```

We renamed the .zip file:

```{r, eval = FALSE}
file.rename(from = "../data/cellphonedb_11_07_2025_010457.zip", to = "../data/cellphonedb.zip") 
```

We reconstructed `cellphonedb.zip` with our changes from Part 1's database harmonisation and provide `cellphonedb.zip` for running CellPhoneDB analysis. If users only want to perform CellPhoneDB analysis on their Seurat object(s), `cellphonedb.zip` would be the selected database for the run. This allows users to compare CellChat's results against CellPhoneDB's inferred results post-harmonisation.

```{r}
rm(interaction_input, gene_input, complex_input, protein_input, transcription_factor_input, uniprot_syn)
```

## Final touches to harmonized databases - format reorganisation

### CCDB

```{r, eval = FALSE}
CCDB <- CC.int %>% 
  relocate(interaction_name, pathway_name, ligand_name, ligand, receptor_name, receptor, LR, complex_ligand, complex_receptor)
```

### CPDB

```{r, eval = FALSE}
unzip("data/cellphonedb.zip", files = c("interaction_table.csv", "multidata_table.csv"), exdir = "data")

multi  <- read_csv("data/multidata_table.csv") %>% select(id_multidata, name)

id_map <- read_csv("data/interaction_table.csv") %>% 
  left_join(multi, by = c("multidata_1_id" = "id_multidata")) %>% 
  mutate(multidata_1_id = ifelse(! is.na(name), name, multidata_1_id)) %>% # change partner_a to uniprot/complex_name
  select(- name) %>% 
  left_join(multi, by = c("multidata_2_id" = "id_multidata")) %>% 
  mutate(multidata_2_id = ifelse(! is.na(name), name, multidata_2_id),     # change partner_b to uniprot/complex_name
         pair = paste(multidata_1_id, multidata_2_id, sep = "|")) %>% 
  select(id_cp_interaction, pair)

CPDB <- CPDB.int %>% 
  mutate(partner_a = case_when(partner_a == "HLAA"  ~ "P04439",
                        partner_a == "HLAB"  ~ "P01889",
                        partner_a == "HLAC"  ~ "P10321",
                        partner_a == "IFNA1" ~ "P01562",
                        .default = partner_a),
         partner_b = ifelse(partner_b == "Glycoprotein_Ib_complex", "GP_complex", partner_b),
         pair = paste(partner_a, partner_b, sep = "|"),
         ligand_name = ifelse(is.na(protein_name_a), partner_a, protein_name_a),
         receptor_name = ifelse(is.na(protein_name_b), partner_b, protein_name_b)) %>% 
  left_join(id_map, by = "pair") %>% 
  select(- id_cp_interaction.x, - pair) %>% 
  relocate(id_cp_interaction = id_cp_interaction.y, classification, ligand_name, ligand, receptor_name, receptor, LR, complex_ligand, complex_receptor)
```

```{r, echo = FALSE}
pacman::p_load(scpeakeR)
```

```{r, eval = FALSE}
CCDB %>% dim()
```

```{r, echo = FALSE}
CCDB %>% dim()
```

```{r, eval = FALSE}
CCDB %>% glimpse()
```

```{r, echo = FALSE}
CCDB %>% glimpse()
```

```{r, eval = FALSE}
CPDB %>% dim()
```

```{r, echo = FALSE}
CPDB %>% dim()
```

```{r, eval = FALSE}
CPDB %>% glimpse()
```

```{r, echo = FALSE}
CPDB %>% glimpse()
```

## Save CCDB and CPDB as R objects to scpeakeR GitHub

```{r, eval = FALSE}
save(CCDB, file = "data/CCDB.rda")
save(CPDB, file = "data/CPDB.rda")
```

This concludes the process performed in **Part 2** of the Database customisation conducted. In **Part 3**, we outline the full convergence of the harmonized databases, providing users with a single merged database comprising of interactions from both CellChat and CellPhoneDB. You may refer to **Part 3** [here](https://nigelhojinker.github.io/scpeakeR/data-raw/Database_harmonisation_part3.html).
