---
title: "scpeakeRDB creation - Part 3"
output: rmarkdown::html_vignette
description: >
  scpeakerDB creation via convergence of CellChatDB and CellPhoneDB
vignette: >
  %\VignetteIndexEntry{scpeakeRDB creation - Part 3}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
date: 2025/01/05
editor_options: 
  chunk_output_type: console
---

# Creating scpeakeRDB

In this document, we outline the process undergone to develop scpeakeRDB, first by identifying interactions not present in each of the databases. We then manually input the mandatory fields required for each package before constructing the databases in a format compatible with their original tools. If you have not viewed "Database customization by scpeakeR - Part 2", you may refer [here]().

## Load required packages & databases

```{r}
pacman::p_load(scpeakeR)
setwd(this.path::here())
rm(list = ls())

data(CCDB, CPDB)
```

## scpeakerDB_CC

We start with the scpeakeRDB-CellChat (scpeakeRDB_CC) database creation. CellChatDB has interactions annotated with pathways in their original structure, and we preserve that same structure for newly added interactions from CellPhoneDB. We made a named mapping vector `pathway_map` with the pathways as the values and the ligands involved in the pathway as the names.  

```{r}
pathway_map <- CCDB %>%
  select(ligand, pathway_name) %>%
  distinct(ligand, .keep_all = TRUE) %>%
  deframe()
```

```{r}
pathway_map %>% length()

pathway_map %>% head()
```

In the next step, we extract interations in CPDB that are not found in CCDB, manually refining the interaction dataframe to resemble more closely to CCDB format:

```{r}
CPDB.only <- CPDB %>%
  filter(LR %in% setdiff(CPDB$LR, CCDB$LR)) %>%
  relocate(interactors) %>%
  mutate(interactors = case_when(
    grepl("^HLA-", interactors)    ~ sub("HLA-", "HLA", interactors),
    grepl("^ERVH48-1", interactors) ~ sub("-", "_", interactors),
    .default = interactors)) %>%
  separate(interactors, into = c("ligand_name", "receptor_name"), sep = "-") %>%
  mutate(ligand_name = ifelse(ligand_name == "ERVH48_1", "ERVH48-1", ligand_name)) %>% # reassign gene to correct name
  mutate(ligand_name   = gsub("\\+", "_", ligand_name),
         receptor_name = gsub("\\+", "_", receptor_name))
```

We not create a `pathway_name` column to `CPDB.only` and map the CellChat-based pathways according to the ligands in CPDB:

```{r}
CPDB.only$pathway_name <- pathway_map[CPDB.only$ligand]
```

```{r}
CPDB.only %>% 
  select(starts_with("ligand"), starts_with("receptor"), pathway_name) %>% 
  head()
```

In most cases, since the ligands in `CPDB.only` are not originally found in CCDB, the interactions they are involved in are not automatically mapped. Here, we manually patched the pathways for the interactions in CPDB:

```{r}
# Keep original copy for comparison
CPDB.only2 <- CPDB.only

# Manual matching of pathway names
CPDB.only <- CPDB.only %>%
  mutate(pathway_name = case_when(
  grepl("^COL", ligand_name) ~ "COLLAGEN",
  grepl("^(WNT5A|WNT5B|WNT11)", ligand_name) ~ "ncWNT",
  grepl("^WNT", ligand_name) ~ "WNT",
  grepl("^INH", ligand_name) ~ "ACTIVIN",
  grepl("^IGFL", ligand_name) ~ "IGF",
  grepl("^CGA_", ligand_name) ~ "CHORIOGONADOTROPIN",
  grepl("^NRXN", receptor_name) ~ "NRXN",
  ligand_name == "PNMT" ~ "Adrenaline",
  ligand_name == "RLN2" ~ "RLN",
  ligand_name == "PODXL2" ~ receptor_name,
  ligand_name == "ANGPT4" ~ "ANGPT",
  ligand_name == "ADM2" ~ "ADM",
  ligand_name == "AKR1B1" ~ "Prostaglandin",
  ligand_name == "NRXN1" & receptor_name == "DAG1" ~ "NRXN",
  receptor_name == "ADGRL1" ~ "ADGRL",
  classification == "Adhesion by Cadherin" ~ "CDH",
  classification == "Adhesion by Fibribogen" ~ "FIBRINOGEN",
  classification == "Adhesion by ICAM" ~ "ICAM",
  classification == "Signaling by Acetylcholine" ~ "Ach",
  classification == "Signaling by TAFA" ~ "TAFA",
  classification == "Signaling by R-spondin" ~ "R-SPONDIN",
  classification == "Signaling by Calsyntenin" ~ "NRXN",
  classification == "Signaling by Agouti-related protein" ~ "MELANOCORTIN",
  classification == "Signaling by BMP" ~ "BMP",
  classification == "Signaling by Interferon" ~ "IFNL",
  is.na(pathway_name) ~ ligand_name,
  .default = pathway_name
))

setdiff(CPDB.only2$pathway_name, CPDB.only$pathway_name) # check for difference in mapping

rm(CPDB.only2, pathway_map)
```

In addition to pathways, CCDB segments their database according to four broad categories of interactions - Secreted Signaling, ECM-Receptor, Cell-Cell Contact and Non-protein Signaling. We constructed an interaction_name column for the newly added interactions by concatenating the ligand-receptor pair and join the pre-exisiting annotations in CCDB via pathway_name. We found 111 interactions which do not have an annotation automatically mapped in the process:

```{r}
CPDB.only <- CPDB.only %>%
  mutate(interaction_name = paste(ligand_name, receptor_name, sep = "_")) %>%
  relocate(interaction_name, pathway_name, ligand_name, ligand, receptor_name, receptor) %>%
  left_join(CCDB %>% select(pathway_name, annotation)) %>%
  distinct(id_cp_interaction, .keep_all = TRUE)

# Extract interactions with unmapped annotation
CPDB.only %>% filter(is.na(annotation)) %>% tabyl(directionality)
# directionality      n    percent
# Adhesion-Adhesion   4    0.03603604
# Ligand-Receptor     102  0.91891892
# Receptor-Receptor   5    0.04504505
```

We manually added these annotations to the newly added interactions, along with an identifer "CPDB:" at the start of these interactions for differentiation:

```{r}
CPDB.only <- CPDB.only %>%
  mutate(annotation = case_when(
    pathway_name == "FIBRINOGEN" ~ "ECM-Receptor",
    pathway_name == "CHORIOGONADOTROPIN" ~ "Non-protein Signaling",
    directionality == "Adhesion-Adhesion" ~ "Cell-Cell Contact",
    directionality == "Receptor-Receptor" ~ "Cell-Cell Contact",
    directionality == "Ligand-Receptor" ~ "Secreted Signaling",
    .default = annotation),
    interaction_name = paste0("CPDB:", interaction_name)) %>%
  distinct(interaction_name, .keep_all = TRUE) # removed one duplicated interaction
```

Now all CellPhoneDB interactions have a pathway_name and a pathway annotation. We begin constructing scpeakeRDB_CC by first recalling the previously revised `CellChatDB.human.new` from Part 2:

### Load CellChatDB.human.new

```{r}
interaction_input <- CellChatDB.human.new$interaction
complex  <- CellChatDB.human.new$complex
cofactor <- CellChatDB.human.new$cofactor
geneInfo <- CellChatDB.human.new$geneInfo
```

```{r}
interaction_input <- bind_rows(interaction_input, CPDB.only %>%
                                 select(interaction_name, pathway_name, 
                                        ligand = ligand_name, receptor = receptor_name, 
                                        annotation, version, LR))
```

Here, we check for any duplicated interaction partners after merging CellChatDB and CellPhoneDB. We observed that duplicated mappings from both databases were due to different Uniprot identifers linked to the same gene symbol. For example, CellChatDB linked *NRXN1* to NRX1B_HUMAN (UniProtKB: P58400) while CellPhoneDB linked *NRXN1* to NRX1A_HUMAN (UniProtKB: QULB1). 

```{r}
interaction_input <- interaction_input %>% 
  mutate(partners = paste(ligand, receptor, sep = "|")) 

interaction_input %>% 
  count(partners) %>% 
  filter(n > 1)
```

As scRNA-seq expression matrix from droplet-based sequencing is not isoform-aware, we compressed these interactions to be the same. In the other case of the gene *PRNP*, CellChatDB linked *PRNP* to PRIO_HUMAN (UniProtKB: P04156) while CellPhoneDB linked *PRNP* to APRIO_HUMAN (UniProtKB: F7VJQ1) the alternative prion protein produced by the same bicistronic gene. Since the expression matrix does not distinguish between the two protein from the same mRNA transcript, we adopted the same collapsing strategy as *NRXN1*.

```{r}
interaction_input <- interaction_input %>% 
  distinct(partners, .keep_all = TRUE) %>% 
  select(-partners) # 3901
```

```{r}
CP_simple_ligand <- CPDB.only %>%
  filter(! complex_ligand) %>%
  pull(ligand_name)

CP_simple_receptor <- CPDB.only %>%
  filter(! complex_receptor) %>%
  pull(receptor_name)

CP_simple <- union(CP_simple_ligand, CP_simple_receptor) # 424

CP_simple[which(! CP_simple %in% geneInfo$Symbol)]
# [1] "C10orf99" "HLAF"     "HLAE"

genes.add <- data.frame(EntryID.uniprot = c("Q6UWK7", "P30511", "P13747"),
                        Symbol = c("C10orf99", "HLAF", "HLAE"))

geneInfo <- bind_rows(geneInfo, genes.add) %>%
  mutate(EntryID.uniprot = ifelse(Symbol == "CGB3", "P0DN86", EntryID.uniprot)) # 26,830

rm(CP_simple, CP_simple_ligand, CP_simple_receptor, genes.add)
```

```{r}
CP_complex_ligand <- CPDB.only %>%
  filter(complex_ligand) %>%
  pull(ligand_name)

CP_complex_receptor <- CPDB.only %>%
  filter(complex_receptor) %>%
  pull(receptor_name)

CP_complex <- union(CP_complex_ligand, CP_complex_receptor) # 120

complex <- complex %>%
  rownames_to_column("complex_name") %>%
  filter(complex_name != "KLRK1_HCST") # unused complex in CellChat

complex[complex == ""] <- NA

complex <- complex %>%
  rowwise() %>%
  mutate(symbol = paste(sort(na.omit(c(subunit_1, subunit_2, subunit_3, subunit_4, subunit_5))), collapse = "_"))

new_complex <- CP_complex[which(! CP_complex %in% complex$symbol)] # 53
```

```{r}
#### Check if genes in "new_complex" found in geneInfo
new.genes <- paste(new_complex, collapse = "_") %>%
  str_split_1("_")

new.genes[which(! new.genes %in% geneInfo$Symbol)] # 0

present_diff_name <- setdiff(CP_complex[which(! CP_complex %in% complex$complex_name)], new_complex) # 17 interactions already in CCDB but diff name

# Remove symbol which requires special handling
present_diff_name <- setdiff(present_diff_name, "ALOX5_ALOX5AP_LTC4S")

map <- complex %>%
  filter(symbol %in% present_diff_name) %>%
  select(symbol, complex_name)

rm(CP_complex, CP_complex_ligand, CP_complex_receptor, present_diff_name, new.genes)
```

```{r}
### Rewrite names of interaction partners to those already present in CellChatDB
interaction_input <- left_join(interaction_input, map, by = c("ligand" = "symbol")) %>%
  mutate(ligand = case_when(
    !is.na(complex_name) ~ complex_name,
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "CYSLTR1" ~ "LXA4-ALOX5",
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "GPR17" ~ "LTC4-LTC4S",
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "OXER1" ~ "5SHpETE_byALOX5",
    .default = ligand
  )) %>%
  select(- complex_name) %>%
  left_join(map, by = c("receptor" = "symbol")) %>%
  mutate(receptor = ifelse(! is.na(complex_name), complex_name, receptor)) %>%
  select(- complex_name)

rownames(interaction_input) <- interaction_input$interaction_name

orig_map <- CCDB %>%
  select(interaction_name, LR)

ccdb_map <- interaction_input %>%
  select(interaction_name, LR) %>%
  left_join(orig_map, by= "interaction_name", suffix = c("_new", "_orig")) %>%
  mutate(LR_new = ifelse(! is.na(LR_orig), LR_orig, LR_new)) %>%
  select(interaction_name, LR = LR_new)

# LR no longer used in interaction_input
interaction_input <- interaction_input %>% 
  select(-LR)

save(ccdb_map, file = "data/ccdb_map.rda")
```

```{r}
### Add new complexes to dataframe
new_complex <- new_complex %>%
  as.data.frame()

colnames(new_complex) <- "complex_name"

new_complex <- new_complex %>%
  separate(complex_name, into = paste0("subunit_", 1:5), sep = "_", fill = "right", remove = FALSE) %>%
  bind_rows(data.frame(complex_name = c("5SHpETE_byALOX5"),
                       subunit_1 = c("LTC4S"),
                       subunit_2 = c("ALOX5"),
                       subunit_3 = c("ALOX5AP")))

complex <- bind_rows(complex, new_complex) %>%
  column_to_rownames("complex_name") %>%
  select(-symbol) # 391

complex[is.na(complex)] <- ""

scpeakerDB_CC <- list()
scpeakerDB_CC$interaction <- interaction_input
scpeakerDB_CC$complex     <- complex
scpeakerDB_CC$cofactor    <- cofactor
scpeakerDB_CC$geneInfo    <- geneInfo

save(scpeakerDB_CC, file = "data/scpeakerDB_CC.rda")

rm(CPDB.only, interaction_input, complex, geneInfo, cofactor, new_complex, map, orig_map)
```

## scpeakeRDB_CP.zip

Now, we curated the scpeakeRDB in zip format that is CellPhoneDB-compatible.

Load the original CellPhoneDB input tables for customization
```{r}
interaction_input <- read_csv("data/interaction_input.csv")
gene_input        <- read_csv("data/gene_input.csv")
complex_input     <- read_csv("data/complex_input.csv")
protein_input     <- read_csv("data/protein_input.csv")
transcription_factor_input <- read_csv("data/transcription_factor_input.csv")
uniprot_syn       <- read_tsv("data/sources/uniprot_synonyms.tsv")
```

Add missing entries to uniprot_syn dataframe:

```{r}
CCDB.only <- CCDB %>%
  filter(LR %in% setdiff(CCDB$LR, CPDB$LR))

CC_simple_ligands <- CCDB.only %>%
  filter(! complex_ligand) %>%
  pull(ligand)

CC_simple_receptors <- CCDB.only %>%
  filter(! complex_receptor) %>%
  pull(receptor)

CC_simple <- union(CC_simple_ligands, CC_simple_receptors)

CC_simple[which(! CC_simple %in% uniprot_syn$Entry)]
# [1] "P16619byCCL3L3" "P01562byIFNA13"

genes.add <- data.frame(Entry = c("P16619byCCL3L3", "P01562byIFNA13"),
                        "Entry Name" = c("CL3L1_HUMAN", "IFNA1_HUMAN"),
                        check.names = FALSE)

uniprot_syn <- bind_rows(uniprot_syn, genes.add)
```

Add current missing entries to gene_input dataframe:

```{r}
genes.add <- data.frame(gene_name = c("CCL3L3", "IFNA13"),
                        uniprot = c("P16619byCCL3L3", "P01562byIFNA13"),
                        hgnc_symbol = c("CCL3L3", "IFNA13"),
                        ensembl = c("ENSG00000276085", "ENSG00000233816"))

gene_input <- bind_rows(gene_input, genes.add)

rm(CC_simple, CC_simple_ligands, CC_simple_receptors, genes.add)
```

Compiling CellChatDB interactions to CellPhoneDB format

```{r}
tmp <- CCDB.only %>%
  select(ligand_name, partner_a = ligand, receptor_name, partner_b = receptor, version, LR) %>%
  mutate(ligand_name = gsub("_", "\\+", ligand_name),
         receptor_name = gsub("_", "\\+", receptor_name),
         interactors = paste(paste0("CCDB:",ligand_name), receptor_name, sep = "-")) %>% # create "interactors" column
  left_join(uniprot_syn %>% select(Entry, `Entry Name`), by = c("partner_a" = "Entry")) %>%
  mutate(protein_name_a = ifelse(! is.na(`Entry Name`), `Entry Name`, NA)) %>%
  select(-`Entry Name`) %>%
  left_join(uniprot_syn %>% select(Entry, `Entry Name`), by = c("partner_b" = "Entry")) %>%
  mutate(protein_name_b = ifelse(! is.na(`Entry Name`), `Entry Name`, NA)) %>%
  select(-`Entry Name`) %>%
  mutate(ligand_name   = gsub("\\+", "_", ligand_name),                   # change format of interaction partners
         receptor_name = gsub("\\+", "_", receptor_name),
         partner_a = ifelse(is.na(protein_name_a), ligand_name,   partner_a),
         partner_b = ifelse(is.na(protein_name_b), receptor_name, partner_b)) %>%
  select(partner_a, partner_b, protein_name_a, protein_name_b, interactors, version, LR)
```

Handle same duplicated interaction in CPDB:
In the earlier section, we observed that some interaction partners were duplicated after database merger due to inconsistent mapping of UniProt identifiers to gene symbols. To deal with these interactions, we compressed the duplicated ligand/receptor to the canonical gene symbol since the expression matrix from scRNA-seq is unable to distinguish between isoforms or protein products from bicistronic genes. In addition, we remove one duplicated interaction we found here:

```{r}
interaction_input <- bind_rows(interaction_input, tmp) %>% 
  filter(! (partner_a == "F7VJQ1" |                                # APRIO_HUMAN
           (partner_a == "Q9ULB1" & partner_b == "O94910") |       # NRX1A_HUMAN
           (partner_a == "Q9ULB1" & partner_b == "Q14118") |       # NRX1A_HUMAN
           (partner_a == "Q8IZI9" & partner_b == "Type_III_IFNR")) # IL28 receptor = Type_III_IFNR
         )                                                         # 3901 interactions

## Extract new complexes to CPDB
CC_complex_ligand <- CCDB.only %>%
  filter(complex_ligand) %>%
  select(complex_name = ligand_name, uniprot = ligand)

CC_complex_receptor <- CCDB.only %>%
  filter(complex_receptor) %>%
  select(complex_name = receptor_name, uniprot = receptor)

CC_complex <- bind_rows(CC_complex_ligand, CC_complex_receptor) %>%
  distinct() # 100

CP_complex <- complex_input %>%
  select(complex_name, starts_with("uniprot_")) %>%
  rowwise() %>%
  mutate(uniprot = paste(sort(na.omit(c(uniprot_1, uniprot_2, uniprot_3, uniprot_4, uniprot_5))), collapse = "_"))

CC_complex <- left_join(CC_complex, CP_complex %>% select(complex_name, uniprot),
                        by = "uniprot", suffix = c("_CC", "_CP")) %>%
  filter(is.na(complex_name_CP) | complex_name_CC != complex_name_CP) # 77

complex_present <- CC_complex %>%
  filter(! is.na(complex_name_CP)) # 48

rm(CC_complex_ligand, CC_complex_receptor, tmp)
```

```{r}
## Rename newly added complexes to original CPDB complex names
interaction_input <- interaction_input %>%
  left_join(complex_present %>% select(complex_name_CC, complex_name_CP), by = c("partner_a" = "complex_name_CC")) %>%
  mutate(partner_a = ifelse(! is.na(complex_name_CP), complex_name_CP, partner_a)) %>%
  select(-complex_name_CP) %>%
  left_join(complex_present %>% select(complex_name_CC, complex_name_CP), by = c("partner_b" = "complex_name_CC")) %>%
  mutate(partner_b = ifelse(! is.na(complex_name_CP), complex_name_CP, partner_b)) %>%
  select(-complex_name_CP)

## Add new complexes to complex_input
complex_new <- anti_join(CC_complex, complex_present) # 29

complex_new <- complex_new %>%
  select(-complex_name_CP) %>%
  rename(complex_name = complex_name_CC) %>%
  separate(uniprot, into = c(paste0("uniprot_", 1:5)), fill = "right", remove = FALSE)

complex_input <- bind_rows(complex_input, complex_new %>% select(complex_name, starts_with("uniprot_"))) %>%
  mutate(version = ifelse(is.na(version), "CellPhoneDBcore==5a", version))

rm(CC_complex, complex_new, complex_present, CP_complex)
```

```{r}
## Extract proteins in complex_input NOT in database
complex_proteins <- complex_input %>%
  rowwise() %>%
  mutate(uniprot = paste(na.omit(c(uniprot_1, uniprot_2, uniprot_3, uniprot_4, uniprot_5)), collapse = "_")) %>%
  pull(uniprot) %>%
  paste(collapse = "_") %>%
  str_split_1(pattern = "_") %>%
  unique()

new_genes    <- complex_proteins[which(! complex_proteins %in% gene_input$uniprot)]    # 7
new_proteins <- complex_proteins[which(! complex_proteins %in% protein_input$uniprot)] # 11

gene_add <- data.frame(gene_name   = c("FREM1", "NPNT", "FRAS1", "FREM2", "SMO", "HSP90AA1", "MMP14"),
                       uniprot = new_genes,
                       hgnc_symbol = c("FREM1", "NPNT", "FRAS1", "FREM2", "SMO", "HSP90AA1", "MMP14"),
                       ensembl     = c("ENSG00000164946", "ENSG00000168743", "ENSG00000138759",
                                       "ENSG00000150893", "ENSG00000128602", "ENSG00000080824",
                                       "ENSG00000157227"))

protein_add <- data.frame(uniprot = new_proteins,
                          protein_name = c("FREM1_HUMAN", "NPNT_HUMAN", "FRAS1_HUMAN", "FREM2_HUMAN", "SMO_HUMAN",
                                           "IL31R_HUMAN", "CNTP1_HUMAN", "CNTN2_HUMAN", "CNTP2_HUMAN", "HS90A_HUMAN",
                                           "MMP14_HUMAN"))
```

```{r}
## Update database with new gene/protein
gene_input    <- bind_rows(gene_input, gene_add)       # 1813
protein_input <- bind_rows(protein_input, protein_add) # 1368

rm(new_genes, new_proteins, gene_add, protein_add, complex_proteins)


## Checking gene_input table
genes <- union( interaction_input$partner_a[! grepl("[: _a-z]", interaction_input$partner_a)],
                interaction_input$partner_b[! grepl("[: _a-z]", interaction_input$partner_b)])

not_included <- setdiff(genes, gene_input$uniprot)

genes.add <- uniprot_syn %>%
  filter(Entry %in% not_included) %>%
  select(gene_name = `Gene Names (primary)`, uniprot = Entry, hgnc_symbol = `Gene Names (primary)`)

gene_input <- bind_rows(gene_input, genes.add)

rm(genes, not_included, genes.add)
```

```{r}
## Adding changes to protein_input table
ligand_protein <- interaction_input %>%
  filter(!is.na(protein_name_a)) %>%
  select(uniprot = partner_a, protein_name = protein_name_a) %>%
  distinct()

receptor_protein <- interaction_input %>%
  filter(!is.na(protein_name_b)) %>%
  select(uniprot = partner_b, protein_name = protein_name_b) %>%
  distinct()

CP_protein <- bind_rows(ligand_protein, receptor_protein)

rm(ligand_protein, receptor_protein)

new_protein <- setdiff(CP_protein$uniprot, protein_input$uniprot)

protein_input <- bind_rows(protein_input, CP_protein %>% filter(uniprot %in% new_protein)) %>%
  distinct()
```

Aligning gene and protien input with interaction_input
```{r}
setdiff(protein_input$uniprot, gene_input$uniprot)    # 0
setdiff(gene_input$uniprot,    protein_input$uniprot) # 194

new_protein <- setdiff(gene_input$uniprot, protein_input$uniprot)

protein_add <- uniprot_syn %>%
  filter(Entry %in% new_protein) %>%
  select(uniprot = Entry, protein_name = `Entry Name`)

setdiff(new_protein, protein_add$uniprot)
# [1] "HLADRA"  "HLADRB5" "HLADRB1" "HLADQA1" "HLADQB1" "HLADQA2" "HLADQB2" "HLADOB"  "HLADMB"  "HLADMA"  "HLADOA"  "HLADPA1" "HLADPB1"

protein_add2 <- data.frame(uniprot = "P05538",
                           protein_name = "DQB2_HUMAN")

protein_add <- bind_rows(protein_add, protein_add2)

protein_input <- bind_rows(protein_input, protein_add)

rm(protein_add, protein_add2, new_protein)

setdiff(protein_input$uniprot, gene_input$uniprot)
# [1] "P05538"

gene_input <- gene_input %>%
  mutate(uniprot = ifelse(gene_name == "HLA-DQB2", "P05538", uniprot))
```

Final edit
```{r}
interaction_input <- interaction_input %>%
  mutate(annotation_strategy = ifelse(is.na(annotation_strategy), "user_curated", annotation_strategy))

orig_map <- CPDB %>%
  mutate(partner = paste(partner_a, partner_b, sep = "|")) %>%
  select(partner, LR)

cpdb_map <- interaction_input %>%
  mutate(partner = paste(partner_a, partner_b, sep = "|")) %>%
  select(partner, LR) %>%
  left_join(orig_map, by = "partner", suffix = c("_new", "_orig")) %>%
  mutate(LR_new = ifelse(! is.na(LR_orig), LR_orig, LR_new)) %>%
  select(partner, LR = LR_new)

save(cpdb_map, file = "data/cpdb_map.rda")

db_map <- full_join(ccdb_map, cpdb_map)

save(db_map, file = "data/db_map.rda")

interaction_input <- interaction_input %>% select(- LR)
```

Create scpeakerDB for CellPhoneDB analysis - scpeakerDB_CP.zip

```{r, eval = FALSE}
if (! dir.exists("scpeaker_cp/data")) sapply(c("scpeaker_cp/data", "scpeaker_cp/data/sources"), dir.create, recursive = TRUE)

### Save input files to data dir
write_csv(interaction_input, file = "scpeaker_cp/data/interaction_input.csv")
write_csv(gene_input,        file = "scpeaker_cp/data/gene_input.csv")
write_csv(complex_input,     file = "scpeaker_cp/data/complex_input.csv")
write_csv(protein_input,     file = "scpeaker_cp/data/protein_input.csv")
write_csv(transcription_factor_input, file = "scpeaker_cp/data/transcription_factor_input.csv")
write_tsv(uniprot_syn, file = "scpeaker_cp/data/sources/uniprot_synonyms.tsv")

## Creating customized CPDB
use_condaenv("scpeaker")

db_create <- import("cellphonedb.utils.db_utils")
db_create$create_db("scpeaker_cp/data")
# Created scpeaker_cp/data\cellphonedb_02_02_2026_133643.zip successfully

file.rename(from = "scpeaker_cp/data/cellphonedb_02_02_2026_133643.zip", to = "scpeaker_cp/data/scpeakerDB_CP.zip") # Saved customized CellPhoneDB .zip file to  CCCtools GitHub
```

```{r, echo = FALSE}
knitr::knit_exit()
```
