---
title: "Database harmonization by scpeakeR - Part 1"
author: "Ho Jin Ker Nigel, Adaikalavan Ramasamy, Gokce Oguz"
description: >
  Database harmonization between CellChatDB.human(v2.2.0) and CellPhoneDB (V5.0.1)
date: 2025/12/30
format: 
  html:
    toc: true
    toc-location: left
    self-contained: true
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

# Introduction

In this document, we outline the database harmonization steps taken to determine the degree of overlap between CellChat and CellPhoneDB, which are listed as:

-   Step 1: Converting gene symbols to Uniprot ID format in CellChatDB
-   Step 2: Identifying duplicates, missing or invalid entries in CellChatDB and CellPhoneDB
-   Step 3: Convert complexes in CellPhoneDB to Uniprot ID format
-   Step 4: Sorting protein subunits of complexes in both databases
-   Step 5: Swapping ligand-receptor pairs to identify missed overlaps
-   Step 6: Partial mapping of complexes

Additionally, we describe the process to customize CellChatDB and CellPhoneDB with the revised changes implemented.

# Load required packages and download CellChatDB and CellPhoneDB databases

## Load required packages

```{r}
pacman::p_load(tidyverse, janitor, CellChat)
setwd(this.path::here())
rm(list = ls())

# File creation
if (! dir.exists("../data/db_utils")) sapply("../data/db_utils", dir.create, recursive = TRUE)

# Function for summarizing overlap between databases  
join_summarize <- function(db1, db2){
  
  db1$LR <- paste0( db1$ligand, "|", db1$receptor )
  db2$LR <- paste0( db2$ligand, "|", db2$receptor )
  
  comb <- inner_join(db1, db2, by = "LR", suffix = c("_db1", "_db2"))
  
  cat(" # ligand-receptors =", comb %>% distinct(LR) %>% nrow(), "\n",
      "# simple ligands    =", comb %>% filter(!complex_ligand_db1)   %>% distinct(ligand_db1)   %>% nrow(), "\n",
      "# simple receptors  =", comb %>% filter(!complex_receptor_db1) %>% distinct(receptor_db1) %>% nrow(), "\n",
      "# complex ligands   =", comb %>% filter(complex_ligand_db1)    %>% distinct(ligand_db1)   %>% nrow(), "\n",
      "# complex receptors =", comb %>% filter(complex_receptor_db1)  %>% distinct(receptor_db1) %>% nrow(), "\n"
  )
  
  return(comb %>% distinct(LR, .keep_all = TRUE))
}
```

NOTE: `join_summarize()` returns a summary table that includes the number of overlapping interactions across two databases, which we define as a match of the ligand\|receptor pair, the number of simple ligands and receptors made up of a single gene, as well as the number of overlapping complexes made up of more than one gene. From CellChat (v2) and CellPhoneDB (v5) onward, both packages included **non-protein signaling** interactions in their latest databases. As such, interactions involving non-protein signaling may contain only one gene, should the hormone/molecule be produced by only one enzyme in its biosynthesis.

# STEP 1: Map Gene Symbol to UniProt in CellChatDB

CellChatDB.human (CCDB) defines ligands and receptors using gene symbols, whereas CellPhoneDB (CPDB) defines them using UniProt IDs. To harmonize the two databases, we used the `geneInfo` slot to map CCDB gene symbols to their corresponding UniProt IDs. For example, the gene symbol EGF was converted to P01133.

CellChat version `r pacman::p_version("CellChat")` stores the database within R under `CellChatDB.human`.

```{r}
pacman::p_version("CellChat")
```

## Prepare CCDB gene-to-uniprot conversion

```{r}
# Gene symbol to UniProt mapping
map <- CellChatDB.human$geneInfo %>% 
  select(Symbol, EntryID.uniprot) %>% 
  drop_na() %>% 
  deframe()

map %>% length()

map %>% head()
```

### Map complex to UniProt complexes and concatenate

CCDB contains complex interaction partners composed of multiple protien subunits. We convert the gene symbols of these proteins to Uniprot ID format.

```{r}
complex <- CellChatDB.human$complex %>% 
  mutate(complex_name = rownames(.))

complex %>% dim()

complex %>% head()

example <- c("12oxoLTB4-PTGR1", "TREM2_TYROBP", "RARA_RXRA_CRABP2", 
             "Pregnenolone-CYP11A1", "HTR3_complex")
```

Before converting the gene symbols to Uniprot ID format, `complex` looks like:

```{r, echo = FALSE}
complex %>% 
  filter(complex_name %in% example) %>%
  knitr::kable()
```

```{r}
# Perfrom gene-to-uniprot conversion
complex$subunit_1 <- map[complex$subunit_1]
complex$subunit_2 <- map[complex$subunit_2]
complex$subunit_3 <- map[complex$subunit_3]
complex$subunit_4 <- map[complex$subunit_4]
complex$subunit_5 <- map[complex$subunit_5]
```

After converting the gene symbols to Uniprot ID format, `complex` now looks like:

```{r, echo = FALSE}
complex %>% 
  filter(complex_name %in% example) %>%
  knitr::kable()
```

We concatenate the Uniprot IDs of the protein subunits for each complex to form a single character value which represents it.

```{r}
complex <- complex %>% select(-complex_name) # remove unused "complex_name" column

complex <- apply(complex, 1, function(x) paste(na.omit(x), collapse = "_"))

complex[which(names(complex) %in% example)]
```

### Convert all interaction partners in CCDB to their respective Uniprot ID format

CCDB stores all `r nrow(CellChatDB.human$interaction)` interactions within R under `CellChatDB.human$interaction`.

```{r}
# Include complex (Uniprot) to map
map <- c(map, complex)

CC.int <- CellChatDB.human$interaction

CC.int %>% dim()
CC.int %>% glimpse()

CC.int$complex_ligand   <- CC.int$ligand   %in% names(complex)
CC.int$complex_receptor <- CC.int$receptor %in% names(complex)

tmp <- join_summarize(CC.int, CC.int)

# Keeping gene/complex name of original database
CC.int$ligand_name   <- CC.int$ligand
CC.int$receptor_name <- CC.int$receptor

# Map Gene Symbol to UniProt in CCDB
CC.int$ligand   <- map[CC.int$ligand]
CC.int$receptor <- map[CC.int$receptor]

CC.int %>% 
  select(interaction_name, pathway_name, ligand_name, ligand, receptor_name, receptor) %>% 
  head()
```

## Download CellPhoneDB database

```{r}
basepath <- "https://raw.github.com/ventolab/cellphonedb-data/refs/heads/master/data"

# Load required files for database customization
CPDB.int <- read_csv(url(file.path(basepath, "interaction_input.csv")))
complex  <- read_csv(url(file.path(basepath, "complex_input.csv")))
```

## Prepare CPDB for database comparison -- Step 1

```{r}
CPDB.int$ligand   <- CPDB.int$partner_a
CPDB.int$receptor <- CPDB.int$partner_b
CPDB.int$complex_ligand   <- CPDB.int$ligand   %in% complex$complex_name
CPDB.int$complex_receptor <- CPDB.int$receptor %in% complex$complex_name

tmp <- join_summarize(CPDB.int, CPDB.int)

rm(map, tmp)

# Step 1 join
step1 <- join_summarize(CC.int, CPDB.int)

rm(step1)
```

**After Step 1 of Database Harmonization, we observed that there were only 585 overlapping interactions across CCDB and CPDB. As expected, no complex interaction partners across both databases overlapped since CPDB complexes were not converted to Uniprot IDs in this step.**

# STEP 2: Addressing duplicated or invalid mappings

## Finding duplicate LR pairs in CCDB after gene-to-uniprot conversion

```{r}
CC.int %>% count(ligand, receptor) %>% filter(n > 1) # 16

CC.int <- CC.int %>% 
  mutate(LR = paste(ligand, receptor, sep = "|"))

dups <- CC.int %>% 
  count(ligand, receptor) %>% 
  filter(n > 1) %>% 
  select(-n) %>% 
  mutate(LR = paste(ligand, receptor, sep = "|"))

left_join( dups, CC.int ) %>% select(ligand_name, ligand, receptor_name, receptor)

rm(dups)
```

### Identifying missing/invalid entries of simple genes

```{r}
CC.int$interaction_name[which(is.na(CC.int$ligand))] 
CC.int$interaction_name[which(is.na(CC.int$receptor))] 
```

We loaded the [uniprotkb](https://www.uniprot.org/) database of only reviewed Uniprot IDs to identify interaction partners not successfully mapped.

```{r}
# Load Uniprot Database
uniprotkb <- read_tsv(url("https://rest.uniprot.org/uniprotkb/stream?download=true&fields=accession%2Cid%2Cprotein_name%2Cgene_names%2Ccc_subcellular_location%2Ccc_pathway%2Clit_pubmed_id%2Cxref_kegg&format=tsv&query=%28*%29+AND+%28reviewed%3Atrue%29+AND+%28model_organism%3A9606%29")) %>% 
  clean_names() %>% 
  filter(! is.na(gene_names)) %>% 
  rowwise() %>% 
  mutate(n_genes = length(str_split_1(gene_names, pattern = " "))) %>% 
  ungroup() %>% 
  separate(gene_names, into = paste0("gene_", 1:max(.$n_genes)), sep = " ") %>% 
  pivot_longer(starts_with("gene"), values_to = "gene") %>% 
  filter(! is.na(gene)) %>% 
  select(-n_genes, -name) %>% 
  relocate(entry, entry_name, gene)

# Checking for non-uniprot entries 
CC.int %>% filter(! complex_ligand,   ! ligand %in% uniprotkb$entry)   %>% pull(ligand,   ligand_name)   # ligand:   CCL3L1, CCL3L3, IFNA1, IFNA13 invalid
CC.int %>% filter(! complex_receptor, ! receptor %in% uniprotkb$entry) %>% pull(receptor, receptor_name) # receptor: GPR1 missing an entry
```

### Manually editing problematic entries

```{r}
tmp <- CC.int %>% 
  mutate(ligand = case_when(ligand_name == "b-Endorphin-POMC" ~ "P01189betaEndo", 
                            ligand_name == "CCL3L1" ~ "P16619",
                            ligand_name == "CCL3L3" ~ "P16619byCCL3L3",
                            ligand_name == "IFNA1"  ~ "P01562",
                            ligand_name == "IFNA13" ~ "P01562byIFNA13",
                            .default = ligand),
         receptor = case_when(grepl("^CALCRL_", receptor_name)   ~ sub("P30988", "Q16602", receptor),
                              interaction_name == "RARRES2_GPR1" ~ "P46091",
                              .default = receptor),
         
         interaction_name = case_when(ligand_name == "IFNA10" ~ "IFNA10_IFNAR1_IFNAR2",
                                      ligand_name == "IFNA13" ~ "IFNA13_IFNAR1_IFNAR2",
                                      .default = interaction_name),
         
         LR = paste(ligand, receptor, sep = "|")) %>% 
  rename(interaction_name_new = interaction_name, ligand_new = ligand, receptor_new = receptor)

# Apply changes to CCDB 
CC.int <- tmp %>% 
  rename_with(~ sub("_new", "", .), ends_with("_new"))
```

## Finding duplicate LR pairs in CPDB after gene-to-uniprot conversion

### CellPhoneDB does not have any duplicated interactions at this stage

```{r}
CPDB.int %>% count(ligand, receptor) %>% filter(n > 1) # 0 
```

### Checking for invalid entries

```{r}
CPDB.int %>% filter(! complex_ligand,   ! ligand %in% uniprotkb$entry)   %>% pull(ligand)   # ligand: HLAA, HLAB, HLAC, IFNA1 not uniprot ID format 
CPDB.int %>% filter(! complex_receptor, ! receptor %in% uniprotkb$entry) %>% pull(receptor) # receptor: no invalid uniprot IDs

rm(uniprotkb)

CPDB.int <- CPDB.int %>% 
  mutate(ligand = case_when(ligand == "HLAA"  ~ "P04439",
                            ligand == "HLAB"  ~ "P01889",
                            ligand == "HLAC"  ~ "P10321",
                            ligand == "IFNA1" ~ "P01562",
                            .default = ligand))
```

```{r}
# Step 2 join
step2 <- join_summarize(CC.int, CPDB.int)

rm(diff, tmp, step2)
```

**After Step 2 of Database Harmonization, we observed that there was a slight increase in the number of overlapping interactions (585 to 591) across CCDB and CPDB.**

# STEP 3: Change nomenclature for CellPhoneDB

In Step 3, we convert the **CPDB** native complex names to their respective Uniprot ID concatenated character value, mirroring Step 1 for **CCDB**.

```{r}
example <- c("12oxoLeukotrieneB4_byPTGR1", "TREM2_receptor", "RAreceptor_RARA_RXRA",
             "Pregnenolone_byCYP11A1", "5HTR3_complex")
```

Before concatenating the Uniprot IDs of the protein subunits, `complex` looks like:

```{r, echo = FALSE}
complex %>% 
  filter(complex_name %in% example) %>%
  select(complex_name, starts_with("uniprot_")) %>% 
  knitr::kable()
```

We convert the complex names to their respective single character value in Uniprot ID format.

```{r}
complex <- complex %>% 
  column_to_rownames("complex_name") %>% 
  select(starts_with("uniprot"))

# Construct CPDB complex map
complex <- apply(complex, 1, function(x) paste(na.omit(x), collapse = "_"))

complex[which(names(complex) %in% example)]

CPDB.int$ligand   <- ifelse(CPDB.int$complex_ligand,   complex[CPDB.int$ligand],   CPDB.int$ligand)
CPDB.int$receptor <- ifelse(CPDB.int$complex_receptor, complex[CPDB.int$receptor], CPDB.int$receptor)

CPDB.int$ligand <- ifelse(CPDB.int$partner_a == "b-Endorphin_byPOMC", "P01189betaEndo", CPDB.int$ligand) # match changes done for same interaction in CCDB

# Step 3 join
step3 <- join_summarize(CC.int, CPDB.int)

rm(complex, step3)
```

# STEP 4: Sort subunits

We observed that some complexes in CCDB and CPDB contained the same protein subunits, but the order in which they were tabulated differed:

### Examples

```{r}
CC.int %>% filter(ligand_name == "IL27 complex") %>% pull(ligand)
CPDB.int %>% filter(partner_a == "IL27") %>% pull(ligand)
```

We sorted the protein subunits making up the complexes in both database

```{r}
sort_paste <- function(x) paste(sort(x), collapse = "_")

CC.int$ligand     <- sapply(strsplit(CC.int$ligand,     split = "_" ), sort_paste )
CC.int$receptor   <- sapply(strsplit(CC.int$receptor,   split = "_" ), sort_paste )
CPDB.int$ligand   <- sapply(strsplit(CPDB.int$ligand,   split = "_" ), sort_paste )
CPDB.int$receptor <- sapply(strsplit(CPDB.int$receptor, split = "_" ), sort_paste )

# Finding duplicated interactions once again after sorting of subunits
## CellChatDB
tmp <- join_summarize(CC.int, CC.int)
```

No duplicated LR pairs were observed in CCDB after sorting subunits of complexes.

```{r}
## CellPhoneDB
tmp <- join_summarize(CPDB.int, CPDB.int)
```

As indicated in the warning above, we observed duplicated complex keys after sorting subunits in CPDB:

```{r}
dups <- CPDB.int %>% 
  count(ligand, receptor) %>% 
  filter(n > 1) %>%  
  select(-n) %>% 
  mutate(LR = paste(ligand, receptor, sep = "|"))
```

We manually differentiated these complexes as such:

```{r}
left_join( dups, CPDB.int)

CPDB.int <- CPDB.int %>% mutate(ligand   = ifelse(partner_a == "LeukotrieneC4_byLTC4S", "P20292_P09917_Q16873", ligand),
                                receptor = ifelse(partner_b == "Type_III_IFNR", "Q8IU57_Q08334", receptor))

CC.int <- CC.int %>% mutate(ligand = ifelse(ligand_name == "LTC4-LTC4S", "P20292_P09917_Q16873", ligand)) # Match name for same complex in CPDB

# After addressing duplicated LR pairs in CellPhoneDB
tmp <- join_summarize(CPDB.int, CPDB.int)

rm(dups, tmp, sort_paste)

# Step 4 join
step4 <- join_summarize(CC.int, CPDB.int)
```

**After Step 4, we observed an increase in the number of overlapping ligand-receptor interactions attributed to the sorting process.**

# STEP 5: Swap partners for unmatched CPDB

Apart from internally standardizing the order of protein subunits within complexes, the order of the interaction partners affected the number of overlapping interactions as well. Since we defined interactions as ligand\|receptor pair, we identified cases where interactions in **CCDB** were arranged as "ligand_A\|receptor_B", while the same interaction was arranged as "receptor_B\|ligand_A" in **CPDB**. This is often the case for cell-contact interactions, where directionality is inconsequential.

We first extracted the interactions which did not overlapped in Step 4, and swapped the interaction partners for those unmatched interactions:

```{r}
found <- step4 %>% 
  select(ligand = ligand_db2, receptor = receptor_db2) %>% 
  mutate(found = 1)

CPDB.int2 <- CPDB.int %>% left_join(found) 

# Extracting rows which were previously NOT MATCHED to CellChatDB
to_swap <- which( is.na( CPDB.int2$found ) )

# Performing swap of ligand and receptor for unmatched interactions
for(i in to_swap){
  
  CPDB.int2$ligand[i]   <- CPDB.int$receptor[i]
  CPDB.int2$receptor[i] <- CPDB.int$ligand[i]
  
  CPDB.int2$partner_a[i] <- CPDB.int$partner_b[i]
  CPDB.int2$partner_b[i] <- CPDB.int$partner_a[i]
  
  CPDB.int2$protein_name_a[i] <- CPDB.int$protein_name_b[i]
  CPDB.int2$protein_name_b[i] <- CPDB.int$protein_name_a[i]
  
  CPDB.int2$complex_ligand[i]   <- CPDB.int$complex_receptor[i]
  CPDB.int2$complex_receptor[i] <- CPDB.int$complex_ligand[i]
}

step5 <- join_summarize(CC.int, CPDB.int2)
```

After swapping the interaction partners of the unmatched interactions only, we observed 48 more overlapping interactions.

```{r}
flip <- step5 %>% 
  filter(LR %in% setdiff(step5$LR, step4$LR)) 
```

All swapped interactions were in the "Cell-Cell Contact" category which validates our approach.

```{r}
flip %>% tabyl(annotation) # all are Cell-Cell Contact
```

We now implement the swap to the original CPDB

```{r}
flipped <- flip$interactors # used for customizing CellPhoneDB database later

# Flipping only interactions that mapped after swapping interaction partners

CPDB.int$flip_LR <- paste(CPDB.int$receptor, CPDB.int$ligand, sep = "|")

to_flip <- which(CPDB.int$flip_LR %in% flip$LR)

cols <- c("ligand", "receptor", "partner_a", "partner_b", "protein_name_a", 
          "protein_name_b", "complex_ligand", "complex_receptor")

for (col in cols) {
  CPDB.int[[col]][to_flip] <- CPDB.int2[[col]][to_flip]
}

CPDB.int$flip_LR <- NULL

rm(CPDB.int2, flip, found, to_flip, to_swap, step4, step5)
```

# STEP 6: Addressing partial matches

Within the complex database in both tools, we observed instances of complexes having the similar names but differ in their subunits.

## CCDB complex

```{r}
CC.complex <- CellChatDB.human$complex %>% 
  rownames_to_column("CellChat_complex") %>% 
  select(CellChat_complex) %>% 
  left_join(CC.int %>% select(ligand_name, ligand),     by = c("CellChat_complex" = "ligand_name")) %>% 
  left_join(CC.int %>% select(receptor_name, receptor), by = c("CellChat_complex" = "receptor_name")) %>% 
  distinct() %>% 
  mutate(CellChat_uniprot = ifelse(is.na(ligand), receptor, ligand)) %>% 
  select(-ligand, -receptor) %>% 
  filter(!is.na(CellChat_uniprot)) # some complexes have NAs for subunits
```

## CPDB complex

```{r}
CPDB.complex <- read_csv(url(file.path(basepath, "complex_input.csv"))) %>% 
  select(CellPhoneDB_complex = complex_name) %>% 
  left_join(CPDB.int %>% select(partner_a, ligand),   by = c("CellPhoneDB_complex" = "partner_a")) %>% 
  left_join(CPDB.int %>% select(partner_b, receptor), by = c("CellPhoneDB_complex" = "partner_b")) %>% 
  distinct() %>% 
  mutate(CellPhoneDB_uniprot = ifelse(is.na(ligand), receptor, ligand)) %>% 
  select(-ligand, -receptor)
```

```{r}
found <- CC.complex %>% 
  inner_join(CPDB.complex, by = c("CellChat_uniprot" = "CellPhoneDB_uniprot"))

CC.only <- CC.complex %>% 
  filter(! CellChat_complex %in% found$CellChat_complex)

CPDB.only <- CPDB.complex %>% 
  filter(! CellPhoneDB_complex %in% found$CellPhoneDB_complex) 

rm(CC.complex, CPDB.complex, found)
```

To identify partial matches, we calculated the jaccard index of the complex subunit composition in both databases, defining a partial match as complexes with a jaccard index more than or equal to 0.50.

$$ Jaccard Index = (number of identical subunits) / (total number of shared subunits) $$ Suppose we have a complexes made up of subunitA_subunitB in CCDB and another similar complexes made up of subunitA_subunitB_subunitC. The pairwise jaccard index for these two complexes will be 0.667 (3 d.p.) since they both share subunits A and B, and the number of unique subunits is 3 (subunits A, B and C).

## Finding potential overlapping subunits in complexes across both databases

```{r}
find_jaccard_index <- function(x, y){
  
  x <- str_split_1(x, pattern = "_")
  y <- str_split_1(y, pattern = "_")
  
  return( length(intersect(x, y)) / length(union(x, y)) )
}

partial.match <- expand.grid(CellChat_uniprot    = CC.only$CellChat_uniprot, 
                             CellPhoneDB_uniprot = CPDB.only$CellPhoneDB_uniprot, 
                             stringsAsFactors    = FALSE)

partial.match$jaccard_index <- mapply(FUN = find_jaccard_index, partial.match[ , 1], partial.match[ , 2])

partial.match <- partial.match %>% 
  arrange(desc(jaccard_index)) %>% 
  filter(jaccard_index >= 0.5) %>% 
  left_join(CC.only) %>% 
  left_join(CPDB.only) %>% 
  relocate(sort(colnames(.))) %>% 
  distinct(CellChat_complex, .keep_all = TRUE) %>% 
  filter(CellChat_complex != "ACVR1B_BMPR2") # no common interaction in both CCDB & CPDB; different complex despite high similarity

rm(CC.only, CPDB.only)

```

## Proposing a new complex name for the partially matched complexes

```{r}
partial.match <- partial.match %>% mutate(proposed_name = c("P00352", "P28329_Q96EP9", "P14784_P31785_Q13261", 
                                                            "O60896_Q16602_Q99527", "P36897_P37173_Q04771", "P07359_P13224_P14770_P40197"))

partial.match
```

## Changing to proposed names in respective databases

### CellChatDB

```{r}
CC.int <- CC.int %>% 
  left_join(partial.match %>% select(CellChat_uniprot, proposed_name), by = c("ligand" = "CellChat_uniprot")) %>% 
  mutate(ligand = ifelse(!is.na(proposed_name), proposed_name, ligand)) %>% 
  select(-proposed_name) %>% 
  left_join(partial.match %>% select(CellChat_uniprot, proposed_name), by = c("receptor" = "CellChat_uniprot")) %>% 
  mutate(receptor = ifelse(!is.na(proposed_name), proposed_name, receptor),
         LR = paste(ligand, receptor, sep = "|")) %>% 
  select(-proposed_name) 
```

### CellPhoneDB

```{r}

CPDB.int <- CPDB.int %>% 
  left_join(partial.match %>% select(CellPhoneDB_uniprot, proposed_name), by = c("ligand" = "CellPhoneDB_uniprot")) %>% 
  mutate(ligand = ifelse(!is.na(proposed_name), proposed_name, ligand)) %>% 
  select(-proposed_name) %>% 
  left_join(partial.match %>% select(CellPhoneDB_uniprot, proposed_name), by = c("receptor" = "CellPhoneDB_uniprot")) %>% 
  mutate(receptor = ifelse(!is.na(proposed_name), proposed_name, receptor),
         LR = paste(ligand, receptor, sep = "|")) %>% 
  select(-proposed_name) 

step6 <- join_summarize(CC.int, CPDB.int)

rm(step6)
```

Save the following objects for customizing CCDB and CPDB in Part 2:

```{r, eval = FALSE}
save(flipped,  file = "../data/db_utils/flipped.rda")
save(CC.int,   file = "../data/db_utils/CC_int.rda")
save(CPDB.int, file = "../data/db_utils/CPDB_int.rda")
```

```{r, echo = FALSE}
if (! file.exists("../data/db_utils/flipped.rda"))  save(flipped,  file = "../data/db_utils/flipped.rda")
if (! file.exists("../data/db_utils/CC_int.rda"))   save(CC.int,   file = "../data/db_utils/CC_int.rda")
if (! file.exists("../data/db_utils/CPDB_int.rda")) save(CPDB.int, file = "../data/db_utils/CPDB_int.rda")
```

Finally, we have a overlap of **2,239** interactions across CCDB and CPDB:

```{r, fig.height = 3, fig.width = 7}
ggvenn::ggvenn(list(CellChat = CC.int$LR, CellPhoneDB = CPDB.int$LR))
```

```{r, echo = FALSE}
knitr::knit_exit()
```

This concludes the process performed in **Part 1** of the Database Harmonization conducted. In **Part 2**, we outline the customization of the CellChatDB and CellPhoneDB with the proposed changes above. You may refer to **Part 2** [here]().


