---
title: "scpeakeRDB creation - Part 3"
author: "Ho Jin Ker Nigel, Adaikalavan Ramasamy, Gokce Oguz"
description: >
  scpeakerDB creation via convergence of CellChatDB and CellPhoneDB
date: 2025/01/05
format: 
  html:
    toc: true
    toc-location: left
    self-contained: true
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  message: false
  warning: false
---

# Creating scpeakeRDB

In this document, we outline the process undergone to develop scpeakeRDB, first by identifying interactions not present in each of the databases. We then manually input the mandatory fields required for each package before constructing the databases in a format compatible with their original tools. If you have not viewed "Database customisation by scpeakeR - Part 2", you may refer [here](https://nigelhojinker.github.io/scpeakeR/data-raw/Database_harmonisation_part2.html). To read "Database harmonisation by scpeakeR - Part 1", you may visit the link [here](https://nigelhojinker.github.io/scpeakeR/data-raw/Database_harmonisation_part1.html).

## Load required packages & databases

We first load the package dependencies and the two curated source databases used throughout this harmonisation workflow. Clearing the workspace helps ensure that subsequent objects are created only from the inputs shown in this document.

```{r}
pacman::p_load(scpeakeR)
setwd(this.path::here())
rm(list = ls())

data(CCDB, CPDB)
```

# scpeakerDB_CC

We start with the scpeakeRDB-CellChat (scpeakeRDB_CC) database creation. CellChatDB has interactions annotated with pathways in their original structure, and we preserve that same structure for newly added interactions from CellPhoneDB. We made a named mapping vector `pathway_map` with the pathways as the values and the ligands involved in the pathway as the names.

```{r}
pathway_map <- CCDB %>%
  select(ligand, pathway_name) %>%
  distinct(ligand, .keep_all = TRUE) %>%
  deframe()
```

## Process CPDB-only interactions to CCDB format

As a quick sanity check, we inspect the size and first few entries of the ligand-to-pathway mapping. This confirms that the mapping vector was created as expected **before** it is applied to CellPhoneDB-only interactions.

```{r}
pathway_map %>% length() # 799

pathway_map %>% head()
```

In the next step, we extract interactions in CPDB that are not found in CCDB, manually refining the interaction dataframe so that it more closely resembles the CCDB format:

```{r}
CPDB.only <- CPDB %>%
  filter(LR %in% setdiff(CPDB$LR, CCDB$LR)) %>%
  relocate(interactors) %>%
  mutate(interactors = case_when(
    grepl("^HLA-", interactors)     ~ sub("HLA-", "HLA", interactors),
    grepl("^ERVH48-1", interactors) ~ sub("-", "_", interactors),
    .default = interactors)) %>%
  separate(interactors, into = c("ligand_name", "receptor_name"), sep = "-") %>%
  mutate(ligand_name = ifelse(ligand_name == "ERVH48_1", "ERVH48-1", ligand_name)) %>% # reassign gene to correct name
  mutate(ligand_name   = gsub("\\+", "_", ligand_name),
         receptor_name = gsub("\\+", "_", receptor_name))
```

We now create a `pathway_name` column in `CPDB.only` and map the CellChat-based pathways according to the ligands in CPDB:

```{r}
CPDB.only$pathway_name <- pathway_map[CPDB.only$ligand]
```

We then preview the relevant ligand, receptor, and pathway columns to confirm that pathway information has been added in the expected format.

```{r}
CPDB.only %>% 
  select(starts_with("ligand"), starts_with("receptor"), pathway_name) %>% 
  head()
```

In most cases, since the ligands in `CPDB.only` are not originally found in CCDB, the interactions they are involved in are not automatically mapped. Here, we manually patched the pathways for the interactions in CPDB:

```{r}
# Keep original copy for comparison
CPDB.only2 <- CPDB.only

# Manual matching of pathway names
CPDB.only <- CPDB.only %>%
  mutate(pathway_name = case_when(
  grepl("^COL", ligand_name) ~ "COLLAGEN",
  grepl("^(WNT5A|WNT5B|WNT11)", ligand_name) ~ "ncWNT",
  grepl("^WNT", ligand_name) ~ "WNT",
  grepl("^INH", ligand_name) ~ "ACTIVIN",
  grepl("^IGFL", ligand_name) ~ "IGF",
  grepl("^CGA_", ligand_name) ~ "CHORIOGONADOTROPIN",
  grepl("^NRXN", receptor_name) ~ "NRXN",
  ligand_name == "PNMT" ~ "Adrenaline",
  ligand_name == "RLN2" ~ "RLN",
  ligand_name == "PODXL2" ~ receptor_name,
  ligand_name == "ANGPT4" ~ "ANGPT",
  ligand_name == "ADM2" ~ "ADM",
  ligand_name == "AKR1B1" ~ "Prostaglandin",
  ligand_name == "NRXN1" & receptor_name == "DAG1" ~ "NRXN",
  receptor_name == "ADGRL1" ~ "ADGRL",
  classification == "Adhesion by Cadherin" ~ "CDH",
  classification == "Adhesion by Fibribogen" ~ "FIBRINOGEN",
  classification == "Adhesion by ICAM" ~ "ICAM",
  classification == "Signaling by Acetylcholine" ~ "Ach",
  classification == "Signaling by TAFA" ~ "TAFA",
  classification == "Signaling by R-spondin" ~ "R-SPONDIN",
  classification == "Signaling by Calsyntenin" ~ "NRXN",
  classification == "Signaling by Agouti-related protein" ~ "MELANOCORTIN",
  classification == "Signaling by BMP" ~ "BMP",
  classification == "Signaling by Interferon" ~ "IFNL",
  is.na(pathway_name) ~ ligand_name,
  .default = pathway_name
))
```

View newly constructed CPDB with pathway names:

```{r}
CPDB.only %>% 
  select(starts_with("ligand"), starts_with("receptor"), pathway_name) %>% 
  head()
```

```{r}
setdiff(CPDB.only2$pathway_name, CPDB.only$pathway_name) # check for difference in mapping (only missing entries)

rm(CPDB.only2, pathway_map)
```

## Assign signalling categories to CPDB-only interactions

In addition to pathways, CCDB segments its database according to four broad categories of interactions: Secreted Signaling, ECM-Receptor, Cell-Cell Contact and Non-protein Signaling. We construct an `interaction_name` column for the newly added interactions by concatenating the ligand-receptor pair and joining the pre-existing annotations in CCDB via `pathway_name`. We found 111 interactions which do not have an annotation automatically mapped in the process:

```{r}
CPDB.only <- CPDB.only %>%
  mutate(interaction_name = paste(ligand_name, receptor_name, sep = "_")) %>%
  relocate(interaction_name, pathway_name, ligand_name, ligand, receptor_name, receptor) %>%
  left_join(CCDB %>% select(pathway_name, annotation)) %>%
  distinct(id_cp_interaction, .keep_all = TRUE)

CPDB.only %>% 
  filter(is.na(annotation)) %>% # Extract interactions with unmapped annotation
  tabyl(directionality)
# directionality      n    percent
# Adhesion-Adhesion   4    0.03603604
# Ligand-Receptor     102  0.91891892
# Receptor-Receptor   5    0.04504505
```

We manually add these annotations to the newly added interactions, along with an identifier "CPDB:" at the start of these interactions for differentiation:

```{r}
CPDB.only <- CPDB.only %>%
  mutate(annotation = case_when(
    pathway_name   == "FIBRINOGEN" ~ "ECM-Receptor",
    pathway_name   == "CHORIOGONADOTROPIN" ~ "Non-protein Signaling",
    directionality == "Adhesion-Adhesion"  ~ "Cell-Cell Contact",
    directionality == "Receptor-Receptor"  ~ "Cell-Cell Contact",
    directionality == "Ligand-Receptor"    ~ "Secreted Signaling",
    .default = annotation),
    interaction_name = paste0("CPDB:", interaction_name)) %>%
  distinct(interaction_name, .keep_all = TRUE) # removed one duplicated interaction

CPDB.only
```

Now all CellPhoneDB interactions have a pathway_name and a pathway annotation. We begin constructing scpeakeRDB_CC by first recalling the previously constructed `CellChatDB.human.new` from Part 2 that contains our implemented changes after database harmonisation:

### Load CellChatDB.human.new

```{r}
data(CellChatDB.human.new)
```

The customised CellChat object from Part 2 is separated into its component tables so that each part can be updated independently.

```{r}
interaction_input <- CellChatDB.human.new$interaction
complex  <- CellChatDB.human.new$complex
cofactor <- CellChatDB.human.new$cofactor
geneInfo <- CellChatDB.human.new$geneInfo
```

The newly curated CellPhoneDB-only interactions are appended to the CellChat interaction table using the columns required by CellChat.

```{r}
interaction_input <- bind_rows(interaction_input, 
                               CPDB.only %>%
                                 select(interaction_name,
                                        pathway_name, 
                                        ligand = ligand_name, 
                                        receptor = receptor_name, 
                                        annotation, 
                                        version, 
                                        LR))
```

Here, we check for any duplicated interaction partners after merging CellChatDB and CellPhoneDB. We observed that duplicated mappings from both databases were due to different UniProt identifiers linked to the same gene symbol. For example, CellChatDB linked *NRXN1* to NRX1B_HUMAN (UniProtKB: P58400) while CellPhoneDB linked *NRXN1* to NRX1A_HUMAN (UniProtKB: Q9ULB1).

```{r}
interaction_input <- interaction_input %>% 
  mutate(partners = paste(ligand, receptor, sep = "|")) 

interaction_input %>% 
  count(partners) %>% 
  filter(n > 1)
```

As scRNA-seq expression matrices from droplet-based sequencing are not isoform-aware, we compressed these interactions to be the same. In the other case of the gene *PRNP*, CellChatDB linked *PRNP* to PRIO_HUMAN (UniProtKB: P04156) while CellPhoneDB linked *PRNP* to APRIO_HUMAN (UniProtKB: F7VJQ1), the alternative prion protein produced by the same bicistronic gene. Since the expression matrix does not distinguish between the two proteins from the same mRNA transcript, we adopted the same collapsing strategy as *NRXN1*.

```{r}
interaction_input <- interaction_input %>% 
  distinct(partners, .keep_all = TRUE) %>% 
  select(-partners) # 3901
```

Next, we identify simple ligand and receptor genes from the CellPhoneDB-only interactions and check whether they already exist in CellChat's `geneInfo` table. Missing genes are manually added with their UniProt identifiers.

```{r}
CP_simple_ligand <- CPDB.only %>%
  filter(! complex_ligand) %>%
  pull(ligand_name)

CP_simple_receptor <- CPDB.only %>%
  filter(! complex_receptor) %>%
  pull(receptor_name)

CP_simple <- union(CP_simple_ligand, CP_simple_receptor) # 424

CP_simple[which(! CP_simple %in% geneInfo$Symbol)] # [1] "C10orf99" "HLAF"     "HLAE"

genes.add <- data.frame(EntryID.uniprot = c("Q6UWK7", "P30511", "P13747"),
                        Symbol = c("C10orf99", "HLAF", "HLAE"))

geneInfo <- bind_rows(geneInfo, genes.add) %>%
  mutate(EntryID.uniprot = ifelse(Symbol == "CGB3", "P0DN86", EntryID.uniprot)) # 26,830

rm(CP_simple, CP_simple_ligand, CP_simple_receptor, genes.add)
```

For complex ligands and receptors, we compare the CellPhoneDB-derived complex definitions against the existing CellChat complex table. Complexes are compared by their sorted subunit symbols so that equivalent complexes can be recognized even when the database-specific names differ.

```{r}
CP_complex_ligand <- CPDB.only %>%
  filter(complex_ligand) %>%
  pull(ligand_name)

CP_complex_receptor <- CPDB.only %>%
  filter(complex_receptor) %>%
  pull(receptor_name)

CP_complex <- union(CP_complex_ligand, CP_complex_receptor) # 120

complex <- complex %>%
  rownames_to_column("complex_name") %>%
  filter(complex_name != "KLRK1_HCST") # unused complex in CellChat

complex[complex == ""] <- NA

complex <- complex %>%
  rowwise() %>%
  mutate(symbol = paste(sort(na.omit(c(subunit_1, subunit_2, subunit_3, subunit_4, subunit_5))), collapse = "_"))

new_complex <- CP_complex[which(! CP_complex %in% complex$symbol)] # 53
```

Before adding new complexes, we verify that all of their component genes are already represented in `geneInfo`. We also create a mapping for complexes that are present in CellChat but named differently from CellPhoneDB.

```{r}
new.genes <- paste(new_complex, collapse = "_") %>%
  str_split_1("_")

new.genes[which(! new.genes %in% geneInfo$Symbol)] # 0

present_diff_name <- setdiff(CP_complex[which(! CP_complex %in% complex$complex_name)], new_complex) # 17 interactions already in CCDB but diff name

# Remove symbol which requires special handling
present_diff_name <- setdiff(present_diff_name, "ALOX5_ALOX5AP_LTC4S")

map <- complex %>%
  filter(symbol %in% present_diff_name) %>%
  select(symbol, complex_name)

rm(CP_complex, CP_complex_ligand, CP_complex_receptor, present_diff_name, new.genes)
```

We use the complex-name mapping to rewrite newly imported interaction partners to the names already used by CellChat. A small number of lipid-related interactions require manual name handling because their complex labels encode different pathway products.

```{r}
# Rewrite names of interaction partners to those already present in CellChatDB
interaction_input <- left_join(interaction_input, map, by = c("ligand" = "symbol")) %>%
  mutate(ligand = case_when(
    !is.na(complex_name) ~ complex_name,
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "CYSLTR1" ~ "LXA4-ALOX5",
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "GPR17" ~ "LTC4-LTC4S",
    ligand == "ALOX5_ALOX5AP_LTC4S" & receptor == "OXER1" ~ "5SHpETE_byALOX5",
    .default = ligand
  )) %>%
  select(- complex_name) %>%
  left_join(map, by = c("receptor" = "symbol")) %>%
  mutate(receptor = ifelse(! is.na(complex_name), complex_name, receptor)) %>%
  select(- complex_name)

rownames(interaction_input) <- interaction_input$interaction_name

orig_map <- CCDB %>%
  select(interaction_name, LR)

ccdb_map <- interaction_input %>%
  select(interaction_name, LR) %>%
  left_join(orig_map, by= "interaction_name", suffix = c("_new", "_orig")) %>%
  mutate(LR_new = ifelse(! is.na(LR_orig), LR_orig, LR_new)) %>%
  select(interaction_name, LR = LR_new)
```

```{r}
ccdb_map %>% head()
```

```{r}
# LR no longer used in interaction_input
interaction_input <- interaction_input %>% 
  select(-LR)
```

The CellChat interaction-to-LR mapping is saved for downstream use when the database is built locally.

```{r, eval = FALSE}
save(ccdb_map, file = "../data/ccdb_map.rda")
```

Finally, new CellPhoneDB-derived complexes are converted into CellChat's complex-table structure and added to the database object alongside the updated interaction and gene information.

```{r}
# Add new complexes to dataframe
new_complex <- new_complex %>%
  as.data.frame()

colnames(new_complex) <- "complex_name"

new_complex <- new_complex %>%
  separate(complex_name, into = paste0("subunit_", 1:5), sep = "_", fill = "right", remove = FALSE) %>%
  bind_rows(data.frame(complex_name = c("5SHpETE_byALOX5"),
                       subunit_1 = c("LTC4S"),
                       subunit_2 = c("ALOX5"),
                       subunit_3 = c("ALOX5AP")))

complex <- bind_rows(complex, new_complex) %>%
  column_to_rownames("complex_name") %>%
  select(-symbol) # 391

complex[is.na(complex)] <- ""

scpeakerDB_CC <- list()
scpeakerDB_CC$interaction <- interaction_input
scpeakerDB_CC$complex     <- complex
scpeakerDB_CC$cofactor    <- cofactor
scpeakerDB_CC$geneInfo    <- geneInfo
```

The complete CellChat-compatible scpeakeR database object can then be saved as an R data file.

```{r, eval = FALSE}
save(scpeakerDB_CC, file = "../data/scpeakerDB_CC.rda")
```

Temporary objects from the CellChat construction step are removed before starting the CellPhoneDB-compatible database build.

```{r}
rm(CPDB.only, interaction_input, complex, geneInfo, cofactor, new_complex, map, orig_map)
```

# scpeakeRDB_CP.zip

Now, we curated the scpeakeRDB in zip format that is CellPhoneDB-compatible.

Load the revised CellPhoneDB from Part 2 (`cellphonedb.zip`) input tables for customisation:

```{r, eval = FALSE}
interaction_input <- read_csv("../data/interaction_input.csv")
gene_input        <- read_csv("../data/gene_input.csv")
complex_input     <- read_csv("../data/complex_input.csv")
protein_input     <- read_csv("../data/protein_input.csv")
transcription_factor_input <- read_csv("../data/transcription_factor_input.csv")
uniprot_syn       <- read_tsv("../data/sources/uniprot_synonyms.tsv")
```

```{r, echo = FALSE}
db_path <- "../data/db_utils/cellphonedb_with_harmonisation_changes/"
interaction_input <- read_csv(file.path(db_path, "interaction_input.csv"))
gene_input        <- read_csv(file.path(db_path, "gene_input.csv"))
complex_input     <- read_csv(file.path(db_path, "complex_input.csv"))
protein_input     <- read_csv(file.path(db_path, "protein_input.csv"))
transcription_factor_input <- read_csv(file.path(db_path, "transcription_factor_input.csv"))
uniprot_syn       <- read_tsv(file.path(db_path, "sources/uniprot_synonyms.tsv"))
```

## Add missing entries to uniprot_syn dataframe:

We isolate CellChat interactions absent from CellPhoneDB and then check which simple protein identifiers are missing from the UniProt synonym table. These missing identifiers are added manually so they can be used in the CellPhoneDB input files.

```{r}
CCDB.only <- CCDB %>%
  filter(LR %in% setdiff(CCDB$LR, CPDB$LR))

CC_simple_ligands <- CCDB.only %>%
  filter(! complex_ligand) %>%
  pull(ligand)

CC_simple_receptors <- CCDB.only %>%
  filter(! complex_receptor) %>%
  pull(receptor)

CC_simple <- union(CC_simple_ligands, CC_simple_receptors)

CC_simple[which(! CC_simple %in% uniprot_syn$Entry)]
# [1] "P16619byCCL3L3" "P01562byIFNA13"

genes.add <- data.frame(Entry        = c("P16619byCCL3L3", "P01562byIFNA13"),
                        "Entry Name" = c("CL3L1_HUMAN", "IFNA1_HUMAN"),
                        check.names = FALSE)

uniprot_syn <- bind_rows(uniprot_syn, genes.add)
```

## Add current missing entries to gene_input dataframe

The same manually curated identifiers are also added to `gene_input`, including their gene symbols and Ensembl identifiers.

```{r}
genes.add <- data.frame(gene_name   = c("CCL3L3", "IFNA13"),
                        uniprot     = c("P16619byCCL3L3", "P01562byIFNA13"),
                        hgnc_symbol = c("CCL3L3", "IFNA13"),
                        ensembl     = c("ENSG00000276085", "ENSG00000233816"))

gene_input <- bind_rows(gene_input, genes.add)

rm(CC_simple, CC_simple_ligands, CC_simple_receptors, genes.add)
```

## Compiling CellChatDB interactions to CellPhoneDB format

CellChat-only interactions are reshaped into the CellPhoneDB interaction schema. During this conversion, partner names are formatted for CellPhoneDB and UniProt synonym information is used to populate protein-name columns where available.

```{r}
tmp <- CCDB.only %>%
  select(ligand_name, partner_a = ligand, receptor_name, partner_b = receptor, version, LR) %>%
  mutate(ligand_name = gsub("_", "\\+", ligand_name),
         receptor_name = gsub("_", "\\+", receptor_name),
         interactors = paste(paste0("CCDB:",ligand_name), receptor_name, sep = "-")) %>% # create "interactors" column
  left_join(uniprot_syn %>% select(Entry, `Entry Name`), by = c("partner_a" = "Entry")) %>%
  mutate(protein_name_a = ifelse(! is.na(`Entry Name`), `Entry Name`, NA)) %>%
  select(-`Entry Name`) %>%
  left_join(uniprot_syn %>% select(Entry, `Entry Name`), by = c("partner_b" = "Entry")) %>%
  mutate(protein_name_b = ifelse(! is.na(`Entry Name`), `Entry Name`, NA)) %>%
  select(-`Entry Name`) %>%
  mutate(ligand_name   = gsub("\\+", "_", ligand_name),                   # change format of interaction partners
         receptor_name = gsub("\\+", "_", receptor_name),
         partner_a = ifelse(is.na(protein_name_a), ligand_name,   partner_a),
         partner_b = ifelse(is.na(protein_name_b), receptor_name, partner_b)) %>%
  select(partner_a, partner_b, protein_name_a, protein_name_b, interactors, version, LR)
```

## Handle same duplicated interaction in CPDB

In the earlier section, we observed that some interaction partners were duplicated after database merger due to inconsistent mapping of UniProt identifiers to gene symbols. To deal with these interactions, we compressed the duplicated ligand/receptor to the canonical gene symbol since the expression matrix from scRNA-seq is unable to distinguish between isoforms or protein products from bicistronic genes. In addition, we remove one duplicated interaction we found here:

After adding the CellChat-only interactions, duplicated or unsupported partner mappings are removed. The remaining complex-containing interactions are then compared against the existing CellPhoneDB complex table to distinguish genuinely new complexes from equivalent complexes that already exist under a different name.

```{r}
interaction_input <- bind_rows(interaction_input, tmp) %>% 
  filter(! (partner_a == "F7VJQ1" |                                # APRIO_HUMAN
           (partner_a == "Q9ULB1" & partner_b == "O94910") |       # NRX1A_HUMAN
           (partner_a == "Q9ULB1" & partner_b == "Q14118") |       # NRX1A_HUMAN
           (partner_a == "Q8IZI9" & partner_b == "Type_III_IFNR")) # IL28 receptor = Type_III_IFNR
         )                                                         # 3901 interactions

## Extract new complexes to CPDB
CC_complex_ligand <- CCDB.only %>%
  filter(complex_ligand) %>%
  select(complex_name = ligand_name, uniprot = ligand)

CC_complex_receptor <- CCDB.only %>%
  filter(complex_receptor) %>%
  select(complex_name = receptor_name, uniprot = receptor)

CC_complex <- bind_rows(CC_complex_ligand, CC_complex_receptor) %>%
  distinct() # 100

CP_complex <- complex_input %>%
  select(complex_name, starts_with("uniprot_")) %>%
  rowwise() %>%
  mutate(uniprot = paste(sort(na.omit(c(uniprot_1, uniprot_2, uniprot_3, uniprot_4, uniprot_5))), collapse = "_"))

CC_complex <- left_join(CC_complex, CP_complex %>% select(complex_name, uniprot),
                        by = "uniprot", suffix = c("_CC", "_CP")) %>%
  filter(is.na(complex_name_CP) | complex_name_CC != complex_name_CP) # 77

complex_present <- CC_complex %>%
  filter(! is.na(complex_name_CP)) # 48

rm(CC_complex_ligand, CC_complex_receptor, tmp)
```

Complex names already present in CellPhoneDB are rewritten to the original CellPhoneDB names, while complexes absent from CellPhoneDB are appended to `complex_input`.

```{r}
# Rename newly added complexes to original CPDB complex names
interaction_input <- interaction_input %>%
  left_join(complex_present %>% select(complex_name_CC, complex_name_CP), by = c("partner_a" = "complex_name_CC")) %>%
  mutate(partner_a = ifelse(! is.na(complex_name_CP), complex_name_CP, partner_a)) %>%
  select(-complex_name_CP) %>%
  left_join(complex_present %>% select(complex_name_CC, complex_name_CP), by = c("partner_b" = "complex_name_CC")) %>%
  mutate(partner_b = ifelse(! is.na(complex_name_CP), complex_name_CP, partner_b)) %>%
  select(-complex_name_CP)

# Add new complexes to complex_input
complex_new <- anti_join(CC_complex, complex_present) # 29

complex_new <- complex_new %>%
  select(-complex_name_CP) %>%
  rename(complex_name = complex_name_CC) %>%
  separate(uniprot, into = c(paste0("uniprot_", 1:5)), fill = "right", remove = FALSE)

complex_input <- bind_rows(complex_input, complex_new %>% select(complex_name, starts_with("uniprot_"))) %>%
  mutate(version = ifelse(is.na(version), "CellPhoneDBcore==5a", version))

rm(CC_complex, complex_new, complex_present, CP_complex)
```

We next identify any proteins from the expanded complex table that are missing from the CellPhoneDB gene or protein tables. These missing entries are manually curated and staged for addition.

```{r}
# Extract proteins in complex_input NOT in database
complex_proteins <- complex_input %>%
  rowwise() %>%
  mutate(uniprot = paste(na.omit(c(uniprot_1, uniprot_2, uniprot_3, uniprot_4, uniprot_5)), collapse = "_")) %>%
  pull(uniprot) %>%
  paste(collapse = "_") %>%
  str_split_1(pattern = "_") %>%
  unique()

new_genes    <- complex_proteins[which(! complex_proteins %in% gene_input$uniprot)]    # 7
new_proteins <- complex_proteins[which(! complex_proteins %in% protein_input$uniprot)] # 11

gene_add <- data.frame(gene_name   = c("FREM1", "NPNT", "FRAS1", "FREM2", "SMO", "HSP90AA1", "MMP14"),
                       uniprot = new_genes,
                       hgnc_symbol = c("FREM1", "NPNT", "FRAS1", "FREM2", "SMO", "HSP90AA1", "MMP14"),
                       ensembl     = c("ENSG00000164946", "ENSG00000168743", "ENSG00000138759",
                                       "ENSG00000150893", "ENSG00000128602", "ENSG00000080824",
                                       "ENSG00000157227"))

protein_add <- data.frame(uniprot = new_proteins,
                          protein_name = c("FREM1_HUMAN", "NPNT_HUMAN",  "FRAS1_HUMAN", 
                                           "FREM2_HUMAN", "SMO_HUMAN",   "IL31R_HUMAN", 
                                           "CNTP1_HUMAN", "CNTN2_HUMAN", "CNTP2_HUMAN", 
                                           "HS90A_HUMAN", "MMP14_HUMAN"))
```

The manually curated gene and protein additions are merged into the CellPhoneDB input tables. We then perform another pass over interaction partners to catch any remaining UniProt identifiers absent from `gene_input`.

```{r}
# Update database with new gene/protein
gene_input    <- bind_rows(gene_input, gene_add)       # 1813
protein_input <- bind_rows(protein_input, protein_add) # 1368

rm(new_genes, new_proteins, gene_add, protein_add, complex_proteins)


# Checking gene_input table
genes <- union( interaction_input$partner_a[! grepl("[: _a-z]", interaction_input$partner_a)],
                interaction_input$partner_b[! grepl("[: _a-z]", interaction_input$partner_b)])

not_included <- setdiff(genes, gene_input$uniprot)

genes.add <- uniprot_syn %>%
  filter(Entry %in% not_included) %>%
  select(gene_name = `Gene Names (primary)`, uniprot = Entry, hgnc_symbol = `Gene Names (primary)`)

gene_input <- bind_rows(gene_input, genes.add)

rm(genes, not_included, genes.add)
```

Protein names from the newly merged interactions are collected from both ligand and receptor sides, then added to `protein_input` where they are not already represented.

```{r}
# Adding changes to protein_input table
ligand_protein <- interaction_input %>%
  filter(!is.na(protein_name_a)) %>%
  select(uniprot = partner_a, protein_name = protein_name_a) %>%
  distinct()

receptor_protein <- interaction_input %>%
  filter(!is.na(protein_name_b)) %>%
  select(uniprot = partner_b, protein_name = protein_name_b) %>%
  distinct()

CP_protein <- bind_rows(ligand_protein, receptor_protein)

rm(ligand_protein, receptor_protein)

new_protein <- setdiff(CP_protein$uniprot, protein_input$uniprot)

protein_input <- bind_rows(protein_input, CP_protein %>% filter(uniprot %in% new_protein)) %>%
  distinct()
```

## Aligning gene and protein input with interaction_input

The gene and protein tables are checked for consistency so that each UniProt identifier used by one table is represented where CellPhoneDB expects it. The remaining HLA-DQB2 discrepancy is handled manually because its database identifier needs to be aligned to `P05538`.

```{r}
setdiff(protein_input$uniprot, gene_input$uniprot)    # 0
setdiff(gene_input$uniprot,    protein_input$uniprot) # 194

new_protein <- setdiff(gene_input$uniprot, protein_input$uniprot)

protein_add <- uniprot_syn %>%
  filter(Entry %in% new_protein) %>%
  select(uniprot = Entry, protein_name = `Entry Name`)

setdiff(new_protein, protein_add$uniprot)
# [1] "HLADRA"  "HLADRB5" "HLADRB1" "HLADQA1" "HLADQB1" "HLADQA2" "HLADQB2" "HLADOB"  "HLADMB"  "HLADMA"  "HLADOA"  "HLADPA1" "HLADPB1"

protein_add2 <- data.frame(uniprot = "P05538",
                           protein_name = "DQB2_HUMAN")

protein_add <- bind_rows(protein_add, protein_add2)

protein_input <- bind_rows(protein_input, protein_add)

rm(protein_add, protein_add2, new_protein)

setdiff(protein_input$uniprot, gene_input$uniprot)
# [1] "P05538"

gene_input <- gene_input %>%
  mutate(uniprot = ifelse(gene_name == "HLA-DQB2", "P05538", uniprot))
```

## Final edit

Before export, newly added interactions are labelled with the user-curated annotation strategy. The CellPhoneDB interaction-to-LR map is also reconstructed and joined with the CellChat map to create a shared lookup table for downstream analyses.

```{r}
interaction_input <- interaction_input %>%
  mutate(annotation_strategy = ifelse(is.na(annotation_strategy), "user_curated", annotation_strategy))

orig_map <- CPDB %>%
  mutate(partner = paste(partner_a, partner_b, sep = "|")) %>%
  select(partner, LR)

cpdb_map <- interaction_input %>%
  mutate(partner = paste(partner_a, partner_b, sep = "|")) %>%
  select(partner, LR) %>%
  left_join(orig_map, by = "partner", suffix = c("_new", "_orig")) %>%
  mutate(LR_new = ifelse(! is.na(LR_orig), LR_orig, LR_new)) %>%
  select(partner, LR = LR_new)

cpdb_map %>% head()
```

The CellPhoneDB interaction-to-LR mapping is saved for downstream use when the database is built locally.

```{r, eval = FALSE}
save(cpdb_map, file = "../data/cpdb_map.rda")
```

```{r}
db_map <- full_join(ccdb_map, cpdb_map)

db_map %>% head()

db_map %>% dim()
```

```{r, eval = FALSE}
save(db_map, file = "../data/db_map.rda")
```

```{r}
interaction_input <- interaction_input %>% select(- LR)
```

## Create scpeakerDB for CellPhoneDB analysis - scpeakerDB_CP.zip

The finalized CellPhoneDB-compatible input tables can be written to a local database directory and passed to CellPhoneDB's database creation utility.

```{r, eval = FALSE}
if (! dir.exists("scpeaker_cp/data")) sapply(c("scpeaker_cp/data", "scpeaker_cp/data/sources"), dir.create, recursive = TRUE)

# Save input files to data dir
write_csv(interaction_input, file = "scpeaker_cp/data/interaction_input.csv")
write_csv(gene_input,        file = "scpeaker_cp/data/gene_input.csv")
write_csv(complex_input,     file = "scpeaker_cp/data/complex_input.csv")
write_csv(protein_input,     file = "scpeaker_cp/data/protein_input.csv")
write_csv(transcription_factor_input, file = "scpeaker_cp/data/transcription_factor_input.csv")
write_tsv(uniprot_syn, file = "scpeaker_cp/data/sources/uniprot_synonyms.tsv")

# Creating customised CPDB
use_condaenv("scpeaker")

db_create <- import("cellphonedb.utils.db_utils")
db_create$create_db("scpeaker_cp/data")
# Created scpeaker_cp/data\cellphonedb_02_02_2026_133643.zip successfully

file.rename(from = "scpeaker_cp/data/cellphonedb_02_02_2026_133643.zip", to = "scpeaker_cp/data/scpeakerDB_CP.zip") # Saved customised CellPhoneDB .zip file to  CCCtools GitHub
```
