---
title: "Demo data for CCCtools package"
output: github_document
date: "2025-08-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, janitor, Seurat)
```


# Download the raw data

We will use the same dataset used in the CellChat tutorial. 
The data is described [here](https://doi.org/10.6084/m9.figshare.13520015.v1).


```{r, eval=FALSE}
pacman::p_load(tidyverse, janitor, Seurat)

load( url("https://figshare.com/ndownloader/files/25950872"), verbose = TRUE )
# Loading objects:
#   data_humanSkin
```
```{r, echo=FALSE, message=FALSE}
fn <- "../data/data_humanSkin.rda"

if( !file.exists(fn) ){
  curl::curl_download("https://figshare.com/ndownloader/files/25950872", 
                    destfile = fn, quiet = TRUE)
}

load(fn)
rm(fn)
```

```{r}
names(data_humanSkin)
```

The metadata shows that there are 4 patients with non-lesional and 4 patients with lesional samples.

```{r}
dim(data_humanSkin$meta)

head(data_humanSkin$meta)

data_humanSkin$meta %>% 
  count(patient.id, condition) %>% 
  arrange(condition, desc(n))

data_humanSkin$meta %>% 
  tabyl(labels, condition)
```

The expression data is normalized counts and consists of 17,328 genes and 7,563 cells.

```{r}
dim(data_humanSkin$data)

data_humanSkin$data[1:5, 1:3]
```

# Convert to Seurat object

As most single-cell data is analyzed using Seurat, we built our functions (i.e. run_cellchat, run_cellphonedb) to use Seurat data as input. We use the `CreateSeuratObject()` function by the `Seurat` package to convert our data to a Seurat object. In most cases, this step is skipped for most users who already have a working Seurat object.

```{r}
seu <- CreateSeuratObject(
  counts    = data_humanSkin$data, 
  meta.data = data_humanSkin$meta,
  min.cells    = 75,       # genes expressed in at least 1% of the samples
  min.features = 200
  )

# Set data matrix to data layer correctly
seu[["RNA"]]$data   <- seu[["RNA"]]$counts 
seu[["RNA"]]$counts <-  NULL

# Cell composition after QC
seu@meta.data %>% tabyl(labels, condition)
```

Let us subset the non-lesional and lesional samples separately:

```{r}
seu.NL <- subset(seu, condition == "NL")
seu.LS <- subset(seu, condition == "LS")

seu.NL@meta.data <- seu.NL@meta.data %>% select(patient.id, labels)

seu.LS@meta.data <- seu.LS@meta.data %>% select(patient.id, labels)

seu.NL
```

```{r}
seu.LS
```

Finally, we write this into the data directory:

```{r, eval = FALSE}
usethis::use_data(seu.NL)
# ✔ Saving "seu.NL" to "data/seu.NL.rda".
# ☐ Document your data (see <https://r-pkgs.org/data.html>).
usethis::use_data(seu.LS)
# ✔ Saving "seu.LS" to "data/seu.LS.rda".
# ☐ Document your data (see <https://r-pkgs.org/data.html>
```

```{r, echo = FALSE}
if( !file.exists("../data/seu.NL.rda") )
  usethis::use_data(seu.NL)

if( !file.exists("../data/seu.LS.rda") )
  usethis::use_data(seu.LS)
```
